# Sampling


## Defining the sample frame

### Defining the distribution area

All above monitoring scopes concern the currently known **distribution area** in the province of Antwerp, 
including the municipalities of Schoten, Brasschaat, and Wuustwezel. 
To define the sample frame, we thus rely on the [Provisional Reference File for Municipal Boundaries](https://www.vlaanderen.be/datavindplaats/catalogus/voorlopig-referentiebestand-gemeentegrenzen-toestand-01-01-2025-correctie-11-02-2025) published by the Flemish government agency "Digital Flanders" (ref). 
This file defines the boundaries of the administrative units in the Flemish Region, 
and provides us with the geometry of the here-relevant Flemish municipalities. 

Alternatively, recent reliable observations could be used to derive the distribution area.


### Defining a buffer around the distribution area

While some monitoring scopes focus on the distribution area, others concern the detection of animals around the distribution area.
Therefore, a buffer area around the distribution area must be defined.
Here, we rely on estimates of the species' mobility / action radius and thus take a buffer of [2000 m]{style="background-color: #E6E6FA;"} width around the distribution area.

[(Recap information on the species' mobility here and adapt width.)]{style="background-color: #E6E6FA;"}


```{r fig-plot-dist-buff-area}
#| include: true
#| message: false
#| fig-cap: "Map of distribution (inner) and buffer (outer) area (both in yellow)." 
#| out-width: 150%



mapview::mapview(
    list(
      maplist_3857$map_antw_merged,
      maplist_3857$map_buff
    ),
    col.regions = list(INBOtheme::vl_darkyellow, INBOtheme::vl_darkyellow),
    alpha.regsions = 0.3,
    color = INBOtheme::vl_darkred
  )

 
```



### Selecting habitat areas in the distribution and buffer area

Within both the distribution and buffer area, we require geometrical information on **(potential) habitat areas** of the Chinese muntjac to define the sample frame.
We know that the species can generally be found in terrestrial habitats with dense vegetation, mainly forests/woodlands, forest edges, hedgerows, abandoned plantations, peri-urban overgrown parks and gardens, and patches of natural vegetation in agricultural areas.
[Based on habitat maps, which habitat type most frequent in Flanders?]{style="background-color: #E6E6FA;"}
We thus use OpenStreetMap data (ref) extracting relevant features via suitable [vegetation](https://wiki.openstreetmap.org/wiki/Vegetation) keys and tags. 

OpenStreetMap key-tag combinations currently used are:

- 'natural = wood' & 'landuse = forest' to obtain (managed and unmanaged) forest ([ref](https://wiki.openstreetmap.org/wiki/Forest))
- 'landuse = orchard' to obtain tree and scrub plantings ([ref](https://wiki.openstreetmap.org/wiki/Tag:landuse%3Dorchard))

[This list can be extended (e.g., by adding key-tag combinations such as 'natural=scrub') or refined (e.g., by adding keys specifying forest type; by considering distance to other features such as cities or open landscape).]{style="background-color: #E6E6FA;"}

To select those (clusters of) habitat areas with a certain minimal size, the information obtained via OpenStreetMap is then processed in the following way:
[(Describe processing here.)]{style="background-color: #E6E6FA;"}
Within the buffer area, habitat areas are additionally required to have a maximal distance of [2000 m]{style="background-color: #E6E6FA;"} (see reasoning above) to the closest habitat area within the distribution area to be selected.

The habitat areas remaining within the distribution area form the **sample frame** for the distribution area and hence monitoring scopes A, B, and D.
[(Add that criterion is minimal overlap; sampling frame can thus extend beyond municipaltity borders.)]{style="background-color: #E6E6FA;"}

For monitoring scope C targeting a buffer around the distribution area, the sample frame is formed by the habitat regions remaining within the buffer area and not selected for the sampling frame of the habitat area. 


```{r fig-plot-sample-frame}
#| include: true
#| message: false
#| fig-cap: "Map of sample frame for distribution (light green) and buffer area (dark green)." 
#| out-width: 150%

mapview::mapview(
    list(
      maplist_3857$map_antw_hab,
      maplist_3857$map_buff_hab
    ),
    col.regions = list(INBOtheme::vl_lightgreen, INBOtheme::vl_darkgreen)
  )

 
```


## Drawing a random sample {.hidden}

HERE. 
Draw a spatially-balanced random sample using a variant of the GRTS-algorithm as implemented in the grtsdb R package.

Steps: 

Tessellation:
With respect to a bounding box overlaying both distribution and buffer area.
Define cell size.
Overlay distribution / buffer area with spatial grid consisting of quadratic cells of dimension $p \times p$ m [here $100 \times 100$ m for illustrational purposes]

Draw balanced sample

Discard grid cells outside sample frame
(minimum overlap)

Keep first N cells.



```{r fig-plot-sample}
#| include: true
#| message: false
#| fig-cap: "Map of sample for distribution area combined (red)." 


source("../grts_sample.R")

mapview::mapview(
    list(
      maplist_3857$map_antw_hab,
      sample_hab_upd
    ),
    col.regions = list(INBOtheme::vl_lightgreen, INBOtheme::inbo_oranjerood),
    alpha.regions = list(0.2, 0.9)
  )

 
```


HERE
For this illustration, we draw a naive random sample of locations.
To this end, we take the following steps:



```{r sampling-create-grid}
#| include: false
#| cache: false
#| results: 'hide'

# create grid over bounding box around distribution area
p = 100
grid_box <- sf::st_make_grid(
  maplist_3857$map_antw_merged,
  cellsize = p # meters, since Belgian Lambert
) |>
  sf::st_sf()
if (FALSE) mapview::mapview(list(maplist_3857$map_antw_merged, grid_box))

# extract grid cells that overlap with habitat polygons
grid_hab <- sf::st_filter(
  grid_box,
  maplist_3857$map_antw_hab
  #,.predicate = sf::st_within
)
if (FALSE) mapview::mapview(list(map_mun_antw_hab, grid_hab))

# extract grid cells that overlap with habitat polygons
# https://github.com/r-spatial/sf/issues/1125


```

2. Draw random sample of $C$ grid cells extracted in the previous step (here $100$ grid cells)

```{r sampling-draw-random-sample}
#| include: false
#| eval: false
#| cache: false
#| results: 'hide'

# --- draw naive random sample ---------------------------------------------

# sample size
ncell <- 500
assertthat::assert_that(ncell < nrow(grid_hab))

# sample
grid_sample <- grid_hab |>
  dplyr::sample_n(tbl = _, size = ncell)

# GRTS
# https://inbo.github.io/grtsdb/articles/basic.html
# implementation?

```



Within each sampled grid cell, a camera trap is assumed to be placed following a certain strategy (e.g., placement in center or corner of grid cell). 

The total number of grid cells associated with the sample frame forms the upper bound of sample sizes to consider in a power / precision analysis.

## Mapping generated data on sampling locations

<!-- HERE: return to assumption on spatial independence (1_samplesize_cd) -->


##### Specifications: Sampling

- Determine grid
  - cell size?
    - [ ] value from literature (250 m; cf. @bollen_gebruik_nodate)
    - [ ] value based on camera angle/radius (usually $< 10$ m?)
    - [x] smallish value to keep things visible
    - [ ] value based on habitat parameters (e.g., forest density; to make sure a grid cell contains suitable areas for camera placement)
  - how to define intersection with habitat areas?
    - [x] grid cell overlaps with habitat areas 
    - [ ] overlap at least $q \gg 0$ %
- Draw a random sample
  - strategy?
    - [x] naive random sample
    - [ ] spatially balanced random sample via the GRTS method (INBO R package *grtsdb*)



