# Appendix

## A

### Relationship between minimal detectable deviation and the Clopper-Pearson confidence interval

<!-- HERE: recap what we are showing here --->

Let $F(k; n, \psi)$ be the cumulative distribution function of the Binomial distribution for $k$ successes - here presences - given $n$ Bernoulli trials with success probability $\psi$:

$$
F(k; n, \psi) = P(X \le k) = \sum_{i=0}^{\lfloor k\rfloor}\binom{n}{i}\psi^i(1-\psi)^{n - i} 
$$
<!-- HERE-->
For zero successes or presences, $k=0$, we get:

\begin{align}
F(k=0; n, \psi) = P(X \le k = 0) &= \sum_{i=0}^{\lfloor k = 0\rfloor}\binom{n}{i}\psi^i(1-\psi)^{n - i} \\
&= \binom{n}{0}\psi^0(1-\psi)^{n - 0} \\
&= (1-\psi)^n
\end{align}

<!-- HERE: check -->
For the null hypothesis of absence, $H_0:\psi = 0$, and type-I error $\alpha = 0.1$, the critical value is 0 
(i.e., the $(1-\alpha)$-quantile of $Binom(\psi = 0, I)$).
Hence, the null hypothesis is rejected if $k > 0$.

Under the alternative hypothesis, $H_a:\psi>0$, we can thus express the type-II error, $\beta$, as:

$$
\beta = F(k=0; n, \psi>0)
$$
In case of $k=0$ the upper bound of the one-sided Clopper-Pearson confidence interval (CI) is (based on [@thulin_cost_2014]):

$$
CI^{upper}_{CP} = 1-\alpha^{1/n}
$$
<!-- HERE: Motivation: Starting from our numerical results... -->
Using this upper bound as the probability of success under the alternative hypothesis in the expression for $\beta$ we get:

\begin{align}
\beta &= F(k=0; n, \psi=CI^{upper}_{CP}) \\
&= F(k=0; n, \psi=1-\alpha^{1/n}) 
\end{align}

Simplifying the equation, we get $\beta = \alpha$:

\begin{align}
\beta &= F(k=0; n, \psi = 1-\alpha^{1/n}) \\
&= (1-(1-\alpha^{1/n}))^n \\
&= (1-1+\alpha^{1/n})^n \\
&= (\alpha^{1/n})^n \\
&= \alpha
\end{align}

Hence, if the alternative probability of success is equal to the upper bound of the one-sided $(1- \alpha)$ Clopper-Pearson CI, 
then the type-I error is the same as the type-II error.
The upper bound of the one-sided $(1- \alpha)$ Clopper-Pearson CI can thus be used as the minimal detectable presence with $(1 - \beta)$ statistical power.

<!-- HERE: alpha does not matter for the critical value, but it matters for the CI ... -->


## B

### Interpretation of the slope paramter in the logistic regression model

Interpretation in terms of the linear predictor, $\eta$:
$\beta_1$ reflects the absolute difference in linear predictor values given a one-unit difference in year*.
Formally, we can write (leaving out subscripts $i$ and $t$ for simplicity but without loss of generality):

\begin{align}
\eta|(year^*_1 + 1) - \eta|year^*_1 &= (\beta_0 + \beta_1 * (year^*_1 + 1)) - (\beta_0 + \beta_1 * year^*_1)\\
\eta|(year^*_1 + 1) - \eta|year^*_1 &= \beta_0 + \beta_1 * year^*_1 + \beta_1 - \beta_0 - \beta_1 * year^*_1\\
\eta|(year^*_1 + 1) - \eta|year^*_1 &= \beta_1 
\end{align}

Interpretation in terms of the odds of presence, $\frac{\psi_{i,t}}{1-\psi_{i,t}}$:
$exp(\beta_1)$ is the factor by which the odds change given a one-unit change in year
(or: $exp(\beta_1)$ reflects the relative difference (ratio) in the odds given a one-unit change in year*).
We can again derive this as (leaving out subscripts):

\begin{align}
log(odds)|(year^*_1 + 1) - log(odds)|year^*_1 &= \beta_1 \quad\quad|\;exp(\cdot) \\
exp(log(odds)|(year^*_1 + 1) - log(odds)|year^*_1) &= exp(\beta_1) \\
\frac{exp(log(odds)|(year^*_1 + 1))}{exp(log(odds)|year^*_1)} &= exp(\beta_1) \\
\frac{odds|(year^*_1 + 1))}{odds|year^*_1} &= exp(\beta_1) \\
odds|(year^*_1 + 1)) &= exp(\beta_1) * odds|year^*_1
\end{align}

<!-- HERE: add interpretation for prop of presence? --->


## C

### Poisson regression

Starting from the distributional assumption in @eq-pois we can use the **Poisson regression model** to linearly decompose $\mu_{i,t}$, the parameter of the Poisson distribution, after applying the log-link function:

$$
log(\mu_{i,t}) = \eta_{i,t} \\
\Leftrightarrow \\
\mu_{i,t} = exp(\eta_{i,t})
$$ {#eq-poisreg1}
with:

- $log(\cdot)$: natural logarithm, here the log-link function
- $\eta_{i,t}$: linear predictor, which is some linear function of a set of predictor variables
- $exp(\cdot)$: exponential (i.e., inverse of log-link) function

The most simple model variant under monitoring scope B (abundance under management - trend) includes an intercept and a linear effect of year, both location unspecific, so index $i$ is dropped:

$$
\eta_{t} = \beta_{0} + \beta_{1} \, year_t^*
$$ {#eq-poisreg2}
with:

- $\beta_{0}$: intercept; linear predictor if $year_t^*= 0$
- $\beta_{1}$: slope associated with $year_t*$, which is some linear function of year

The **parameter of interest** - corresponding to our monitoring target 'trend' - is obviously the **slope**, $\beta_{1}$.
It's interpretation in terms of the linear predictor, $\eta$, is the one we are familiar with from simple linear regression:
$\beta_1$ reflects the absolute change in linear predictor values given a one-unit change in year*.

The interpretation in terms of the expected number of observed animals, $\mu$, needs to incorporate the link-function used:
Here, $exp(\beta_1)$ reflects the relative difference (ratio) in expected numbers of observations given a 1-unit difference in year*.
So, $exp(\beta_1)$ is the factor by which $\mu$ changes given a one-unit change in year
($\mu|(year^*_1 + 1) = exp(\beta_1)*\mu|year^*_1$),
and $(exp(\beta_1)-1)*100$ is the percentage change in $\mu$ associated with a one-unit change in year.

The following figure shows a number of different trends implied by different sizes of $\beta_1$ for both $\eta$ and $\mu$.

```{r fig-size-trend}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Visualization of different effect sizes of beta_1 in the Poisson regression model"
#| out-width: 150%

data_trend <- data.frame(
  # size of linear effect of year
  effect_size = -seq(0, 2, 0.2)
) |>
  tidyr::crossing(
   year_star = seq(0,15)
) |>
  dplyr::mutate(
    eta =  effect_size*year_star + 2,
    mu = exp(eta),
    effect_size = factor(effect_size) |> forcats::fct_rev()
    ) |>
  tidyr::pivot_longer(
    cols = tidyselect::all_of(c("eta", "mu")),
    names_to = "out_name",
    values_to = "out_value"
  )
ggplot2::ggplot(
  data = data_trend, 
  mapping = ggplot2::aes(x = year_star, y = out_value, group = effect_size)
  ) + 
  ggplot2::geom_line(ggplot2::aes(color = effect_size), linewidth = 1.2) +
  ggplot2::facet_wrap(~ out_name, scales = "free_y", ncol = 2) + 
  ggplot2::labs(x = "year*", y = "outcome", color = "size beta1") +
  ggplot2::scale_x_continuous(breaks = data_trend$year_star |> unique()) + 
  ggplot2::scale_colour_viridis_d(direction = -1) + 
  ggplot2::theme_bw()

```

In case $\beta_1 = 0$, we have `r round((exp(0)-1)*100, digits = 0)` %,
if $\beta_1 = -0.2$, we have around `r round((exp(-0.2)-1)*100, digits = 0)` %, 
if $\beta_1 = -0.4$, we have around `r round((exp(-0.4)-1)*100, digits = 0)` %, 
if $\beta_1 = -1$, we have around `r round((exp(-1)-1)*100, digits = 0)` % decrease per year. 

## D

### Simulation-based power analysis using the mlpwr R package

```{r get-mlpwr-res-append}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'


```

```{r fig-mlpwr-app}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Results of the mlpwr-power analysis - simulated vs. surrogate power per condition"
#| out-width: 150%
#| fig-height: 20
#| fig-width: 8

# load results
load("../../../output/sample_design/mlpwr_logreg/gridpars_20251022-133925.Rda")
load("../../../output/sample_design/mlpwr_logreg/res_list_20251022-133925.Rda")
  
# create figs
source("../power_mlpwr_logreg_vis.R")

library(patchwork)

(plot_res_iter_b0large/plot_res_iter_b0small) + 
  patchwork::plot_annotation(tag_levels = 'A')



```
