# Sampling

```{r setup}
#| message: false
#| results: 'hide'
#| cache: true

mapfile <- "../../data/gis/sample/sampleframe_maplist_31370.Rds"
samplingscript <- "../sample_design/grts_sample.R"
source("../sample_design/plot_sample_frame_sample.R")

# figure settings
knitr::opts_chunk$set(
    out.width = "100%", # enough room to breath
    fig.width = 9,     # reasonable size
    fig.asp = 0.618,   # golden ratio
    fig.align = "center" # mostly what I want
)

```

## Defining the sample frame

### Defining the distribution area

All above monitoring scopes concern the currently known **distribution area** in the province of Antwerp, 
HERE
here we limit ourselves to...alternative areas might include the ones followed up in the province of Antwerp in the context of the M&M project (ref section)
including the municipalities of Schoten, Brasschaat, and Wuustwezel.
To define the sample frame, more specifically, the distribution area, we thus rely on the Provisional Reference File for Municipal Boundaries published by the Flemish government agency Digital Flanders [@agentschap_digitaal_vlaanderen_voorlopig_2025]. 
This file defines the boundaries of the administrative units in the Flemish Region, 
and provides us with the geometry of the here-relevant Flemish municipalities. 

Alternative, more data-driven approaches to defining the distribution area are of course possible.
For instance, most recent (verified) observations could be used in combination with a spatial clustering or density estimation approach.

### Defining a buffer around the distribution area

While some monitoring scopes focus on the distribution area itself, others concern the detection of animals around the distribution area.
Therefore, a **buffer area** around the distribution area must be defined.
Here, we rely on estimates of the species' action radius (@sec-biology) and thus take a buffer of 2000 m width around the distribution area.

Figure @fig-plot-dist-buff-area maps the distribution and buffer area onto an OpenStreetMap tile (ref). 

```{r fig-plot-dist-buff-area}
#| include: true
#| message: false
#| results: 'hide'
#| fig-cap: "Map of distribution (inner) and buffer (outer) area (both in yellow)." 
##| out-width: 150%
##| fig-height: 4


plot_dist_buff_0

```

### Selecting habitat areas in the distribution and buffer area

Within both the distribution and buffer area, we require geometrical information on **(potential) habitat areas** of the Chinese muntjac to define the sample frame.
We know that the species can generally be found in terrestrial habitats with dense vegetation, mainly woodlands, forest edges, hedgerows, abandoned plantations, peri-urban overgrown parks and gardens, and patches of natural vegetation in agricultural areas (@sec-biology).
We thus use OpenStreetMap data (ref) extracting relevant features via suitable [vegetation](https://wiki.openstreetmap.org/wiki/Vegetation) keys and tags. 

The OpenStreetMap key-tag combinations currently used are:

- 'natural = wood' & 'landuse = forest' to obtain (managed and unmanaged) forest ([reference](https://wiki.openstreetmap.org/wiki/Forest))
- 'landuse = orchard' to obtain tree and scrub plantings ([reference](https://wiki.openstreetmap.org/wiki/Tag:landuse%3Dorchard))

This list can of course be extended (e.g., by adding key-tag combinations such as 'natural=scrub') or refined (e.g., by adding keys specifying forest type; by considering distance to other features such as cities or open landscape).

In addition, the selected geometries/polygons can further be filtered:
One might for instance 
To select those (clusters of) habitat areas with a certain minimal size, the information obtained via OpenStreetMap is then processed in the following way:
HERE
[(Describe processing here.)]{style="background-color: #E6E6FA;"}
Within the buffer area, habitat areas are additionally required to have a maximal distance of 2000 m (see reasoning above) to the closest habitat area within the distribution area to be selected.

The habitat areas remaining within the distribution area form the **sample frame** for the distribution area and hence monitoring scopes A, B, and D.
HERE
Add that criterion is minimal overlap; sampling frame can thus extend beyond municipaltity borders.

Figure @fig-plot-sample-frame maps an example sample frame...HERE 
For monitoring scope C targeting a buffer around the distribution area, the sample frame is formed by the habitat regions remaining within the buffer area and not selected for the sampling frame of the habitat area.


```{r fig-plot-sample-frame}
#| include: true
#| message: false
#| results: 'hide'
#| fig-cap: "Map of sample frame for distribution (light green) and buffer area (dark green)." 
##| out-width: 150%
##| fig-height: 4

plot_dist_buff
 
```

## Drawing a random sample

Draw a spatially-balanced random sample using a variant of the Generalised Random Tesselation Stratified [GRTS, @olsen_spatially_2012] approach as implemented in the grtsdb R package [@onkelinx_grtsdb_2022].

The GRTS approach to sampling ...[HERE]

The implemented variant [HERE]
It allows to... [link!?]

In terms of the present implementation, the following steps are required:

Tessellation:
With respect to a bounding box overlaying both distribution and buffer area.
Define cell size.
Overlay distribution / buffer area with spatial grid consisting of quadratic cells of dimension $p \times p$ m [here $100 \times 100$ m for illustrational purposes]

Draw balanced sample

Discard grid cells outside sample frame
(minimum overlap)

Keep first N cells.


Figure @fig-plot-sample maps an example sample frame...HERE 



```{r fig-plot-sample}
#| include: true
#| message: false
#| fig-cap: "Map of sample for distribution area combined (red)." 
##| out-width: 150%
##| fig-height: 4

plot_dist_sample
 
```




## Mapping generated data on sampling locations

<!-- HERE: return to assumption on spatial independence (1_samplesize_cd) -->


- Determine grid
  - cell size?
    - [ ] value from literature (250 m; cf. @bollen_gebruik_nodate)
    - [ ] value based on camera angle/radius (usually $< 10$ m?)
    - [x] smallish value to keep things visible
    - [ ] value based on habitat parameters (e.g., forest density; to make sure a grid cell contains suitable areas for camera placement)
  - how to define intersection with habitat areas?
    - [x] grid cell overlaps with habitat areas 
    - [ ] overlap at least $q \gg 0$ %



Neeckx, p.21:
During the placement of the cameras in the field, the exact location deviated a bit from the simulation due to the need for a suitable attachment to a tree, limitation due to vegetation, or accessibility.
