# Sampling design {#sec-sampling}

```{r setup-sampling}
#| message: false
#| results: 'hide'
#| cache: true

mapfile <- "../../data/gis/sample/sampleframe_maplist_31370_smooth.Rds"
mapfile_exampl <- "../../data/gis/sample/sampleframe_maplist_31370.Rds"
mapfile_bg <- "../../data/gis/sample/sampleframe_maplist_31370.Rds"
samplingscript <- "../sample_design/grts_sample.R"
source("../sample_design/plot_sample_frame_sample.R")
source("../sample_design/smooth_sample_frame_example.R")
source("../sample_design/sample_frame_size.R")

# figure settings
knitr::opts_chunk$set(
    out.width = "100%", 
    fig.width = 9,
    fig.asp = 0.618, # golden ratio (portrait)
    fig.align = "center" 
)

```



In this section we discuss the **sampling design** for the monitoring needs presented in @sec-overview-needs.
This includes both the sample frame as well as the sampling procedure.

A **sample frame** can be defined as a collection of all elements in the target population.
In this case, it is thus based on a GIS map layer containing all potential habitat patches (i.e., a collection of polygons along with geographical coordinates) 
in the currently known distribution area of the Chinese muntjac or in a surrounding buffer area [cf. @olsen_spatially_2012; @wouters_ontwerp_2008-1].
Elements of the target population and, by consequence, sample frame are defined by means of tessellation using a grid of square cells, which the habitat map layer is overlaid with. 
Only grid cells overlapping with the habitat polygons are retained.
Sampling locations can then be provisionally equated with these grid cells (but see @sec-methodology).
Conveniently, tessellation does not require any additional steps, but is an integral part of the sampling procedure anyway.

To draw a representative sample of locations from the target population, we rely on random sampling.
More specifically, we draw a **spatially balanced random sample** of locations using the Generalised Random Tessellation Stratified (GRTS) approach [@olsen_spatially_2012; @stevens_spatially_2004].
The approach ensures that the spatial distribution of the sampled locations reflects the distribution of elements of the target population.
Here we rely on a variant of the GRTS approach as implemented in the R package *grtsdb* [@onkelinx_grtsdb_2022].
The implementation is flexible in various ways, and can for instance accommodate changes in the sample frame due to habitat changes over time [@onkelinx_basic_2021].


## Defining the sample frame

### Defining the distribution area

All monitoring scopes concern (in some way) the currently known distribution area in the province of Antwerp.
Here we consider the **municipalities of Schoten, Brasschaat, and Wuustwezel**, but the list may be extended.
To define the distribution area, we thus rely on the Provisional Reference File for Municipal Boundaries published by the Flemish government agency Digital Flanders [@agentschap_digitaal_vlaanderen_voorlopig_2025]. 
This file defines the boundaries of the administrative units in the Flemish Region, 
and provides us with the geometry of the here-relevant municipalities. 

Alternatively, more data-driven approaches to defining the distribution area are of course possible.
For instance, the most recent (verified) observations of Chinese muntjac (in and around Antwerp) could be used in combination with a spatial clustering or density estimation approach.

### Defining a buffer around the distribution area

While some monitoring needs focus on the distribution area itself, others concern the detection of animals around the distribution area.
Therefore, a **buffer area** around the distribution area must be defined.
Here, we rely on estimates of the species' action radius (@sec-biology) and thus expand the distribution area with a buffer distance of 2000 m (using the *st_buffer()* function of the *sf* R package [@pebesma_simple_2018] with default arguments).

@fig-plot-dist-buff-area maps the distribution and buffer area onto an OpenStreetMap tile (ref). 

```{r fig-plot-dist-buff-area}
#| include: true
#| message: false
#| results: 'hide'
#| fig-cap: "Map of distribution (inner) and buffer (outer) area (both in yellow)." 

plot_dist_buff_0

```

### Selecting habitat areas in the distribution and buffer area

Within both the distribution and buffer area, we require geometrical information on **(potential) habitat areas** of the Chinese muntjac to define the sample frame.
We know that the species can generally be found in terrestrial habitats with dense vegetation, mainly woodlands, forest edges, hedgerows, abandoned plantations, peri-urban overgrown parks and gardens, and patches of natural vegetation in agricultural areas (@sec-biology).
We thus use OpenStreetMap data (ref) extracting relevant features via suitable [vegetation](https://wiki.openstreetmap.org/wiki/Vegetation) keys and tags. 

The OpenStreetMap key-tag combinations currently used are:

- "natural = wood" & "landuse = forest" to obtain (managed and unmanaged) forest ([reference](https://wiki.openstreetmap.org/wiki/Forest))
- "landuse = orchard" to obtain tree and scrub plantings ([reference](https://wiki.openstreetmap.org/wiki/Tag:landuse%3Dorchard))

This list can of course be extended (e.g., by adding key-tag combinations such as "natural=scrub") or refined (e.g., by adding keys specifying forest type; by considering distance to other features such as cities or open landscape).

In addition, the selected polygons can be processed further.
To select those (clusters of) habitat areas with a certain minimal size, the information obtained via OpenStreetMap is then processed in the following way:
For instance, highly fragmented (parts of) habitat patches can be removed from the sample frame by a procedure referred to as "opening" in image processing: 
It is morphological operation that consists of an "erosion" followed by a "dilation" step. 
An erosion shrinks a polygon by moving its edges inward,
while dilation expands it again by pushing the edges outward (i.e., we already used this earlier when defining a buffer around the distribution area). 
Applied in sequence, opening removes small bumps or thin protrusions while keeping the main shape of the polygon intact.
@fig-plot-ploy-smooth illustrates the procedure for an exemplary polygon of the sample frame (Park Vordenstein and adjacent habitat patches).
We again rely on the *st_buffer()* function of the *sf* R package with default arguments. 
In @fig-plot-ploy-smooth a relatively large buffer difference (i.e., 100 m) is used during both erosion and dilation, for illustrational purposes.
For the actual sample frame, we employ a smaller distance of 20 m.


```{r fig-plot-ploy-smooth}
#| include: true
#| message: false
#| results: 'hide'
#| fig-cap: "Opening a polygon from the sample frame." 

plot_poly_smooth

```

The habitat areas remaining within the distribution area form the **sample frame** for the monitoring needs
"confirm population decline under management", "confirm eradication under management", and "detect presence after management".
For monitoring need "detect spread under management" targeting a buffer around the distribution area, the sample frame is formed by the habitat regions remaining within the buffer area and not selected for the sample frame of the habitat area.
Figure @fig-plot-sample-frame maps these sample frames.


```{r fig-plot-sample-frame}
#| include: true
#| message: false
#| results: 'hide'
#| fig-cap: "Sample frame for distribution (light green) and buffer area (dark green)." 
##| out-width: 150%
##| fig-height: 4

plot_dist_buff
 
```

In terms of total size we have:

- sample frame distribution area: `r maplist_31370_area$map_antw_hab$area_ha |> as.numeric() |> round(digits = 2)` hectares
- sample frame buffer area: `r maplist_31370_area$map_buff_hab$area_ha |> as.numeric() |> round(digits = 2)` hectares


## Drawing a spatially balanced random sample

We draw a spatially-balanced random sample using a variant of the GRTS approach as implemented in the *grtsdb* R package [@onkelinx_grtsdb_2022].
For the tessellation, we define a grid cell size of 100 m^2^ for illustrational purposes (i.e., so that sampled grid cells remain visible).
A possible value for the actual sampling could be 10 m^2^, as 10 m seem a liberal upper limit for a detection distance of camera traps [@sec-monitoring-methods; @neeckx_occurrence_2024 estimated distances of around 5 m].
The total number of grid cells of this size per sample frame would be:

- sample frame distribution area: `r maplist_31370_grid_size$map_antw_hab`
- sample frame buffer area: `r maplist_31370_grid_size$map_buff_hab`


Figure @fig-plot-sample maps a sample of 100 locations from the sample frame of the distribution area.



```{r fig-plot-sample}
#| include: true
#| message: false
#| fig-cap: "Sampling locations for distribution area (red)" 
##| out-width: 150%
##| fig-height: 4

plot_dist_sample
 
```

