# Sample size planning - scopes C & D {#sec-scope-cd}

## Data-generating model

For these scopes, we stick to certain **data-generating distributions** as the data-generating models.
In @sec-scope-a, we combine data-generating distributions with regression modelling to further decompose the parameters of the data-generating distribution.

### Data-generating distribution

#### Bernoulli and Binomial distribution {#sec-bernbinom}

Monitoring scopes C and D concern detection and thus the presence/absence of animals as outcomes. 
We can rely on the **Bernoulli distribution** as the data-generating distribution. 
The parameter of focus is the probability of success, here the probability of presence:

$$
obs_{i,t=t'} \sim Bern(\psi_{i,t=t'})
$$ {#eq-bern1}

- $obs_{i,t=t'}  \in \{0,1\}$: observation of animals at location $i$, year $t'$
  - aggregated over days: at least one animal on at least one camera frame on at least one day
- $\psi_{i,t=t'}  \in [0,1]$: probability of presence at location $i$, year $t'$
  - matching above aggregation (i.e., it concerns the presence of at least one animal on at least ...) 

We can think of the probability of presence featuring in @eq-bern1 as being an observed one, which should be linked to a true probability of presence.
Ideally the link is a one-to-one link.
An imperfect detection probability of the monitoring method can weaken the link. 
For now we assume a camera trap detection probability of one, hence we can perfectly detect animals if they are present.
<!--The same animals being present multiple times or multiple animals being present is accommodated by the above 'aggregated interpretation' of the probability of presence.
The same animals being present multiple times or multiple animals being present should not be an issue and can be accommodated by thinking of $\psi$ as the probability of at least one animal being present.-->

If we assume that the probability of presence is **invariant across locations** (i.e., $\psi_{i,t=t'} = \psi_{t=t'}$),
it follows from above that the number of locations with animal sightings (i.e., $nobs_{t=t'} := \sum_{i=1}^I obs_{i,t=t'}$) can be thought of as realisations of the **Binomial distribution** in function of $\psi_{t=t'}$:

$$
nobs_{t=t'} \sim Binom(\psi_{t=t'}, I)
$$ {#eq-binom}

- $nobs_{t=t'}  \in \{0,1,2,3, ..., I\}$: number of observations of animals summed across all locations, at year $t'$
- $\psi_{t=t'}  \in [0,1]$: probability of presence at time $t = t'$
  - here assumed invariant across all locations
- $I$: the total number of locations

If we allow for **location-specific probabilities of presence**, the situation gets slightly more complicated.
In that case $nobs_{t=t'}$ could have been generated by the **Poisson Binomial distribution**, which captures a set of Bernoulli trials with unequal probabilities of success/presence.
There is no particular assumption made about how the different probabilities of presence arise (i.e., they are assumed to be fixed in statistical terms).
Importantly, the distribution can be approximated by a Binomial distribution, with a probability of presence corresponding to the mean probability of presence across locations (see @fig-plot-poisbinom).
This means that the following results, which hold exactly for the location-unspecific case, hold approximately for the location-specific case.
<!-- HERE quantification of approximation error?-->
<!-- HERE plot 3 different values of psi (e.g., .1, .5, .9) -->
<!-- Beta Binomial distribution: https://stats.stackexchange.com/questions/390023/poisson-binomial-vs-beta-binomial]-->


```{r fig-plot-poisbinom}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Estimated densities of the Poisson binomial (solid) vs. approximating Binomial distribution (dashed)"
#| out-width: 150%
#| fig-height: 5
#| fig-width: 10

# function to draws from the poisson binomial distribution
source("../_functions/rpoisbinom.R")

# number of locations
I = 100

data_poisbin <- lapply(
  seq(1,5000),
  \(i){
    rpoisbinom(
      I = I,
      p_min = 0, p_max = 0.5,
      runif_seed = 123, rbinom_seed = i
      ) |>
      cbind(i = i)
  }
) |> dplyr::bind_rows()

# density poisson binomial
tmp <- data_poisbin$y_sum |> density()
dens_poisbin <- data.frame(
  x = tmp$x,
  y = tmp$y
)

# density binomial
dens_binom <- lapply(
  seq(0,I),
  \(i){
    data.frame(
      x = i,
      y = dbinom(x = i, size = I, prob = data_poisbin$psi_mean[1])
    )
  }
) |> dplyr::bind_rows()

ggplot2::ggplot(data = dens_binom, ggplot2::aes(x = x, y = y)) + 
  ggplot2::geom_line(linetype = "dashed") + 
  ggplot2::geom_line(data = dens_poisbin) + 
  ggplot2::theme_bw() + 
  ggplot2::labs(
    x = "Number of successes",
    y = "Probability"
  )

```

Finally, one assumption should be highlighted: 
Under the binomial distribution there are no dependencies between locations.
This is a defensible assumption if camera traps are located relatively far apart, especially if a spatially balanced sample is taken.
The assumption becomes more questionable if the sample size increases such that camera traps stand closer together:
in that case, we might expect positive dependencies if two camera traps are within one territory of one individual, 
and negative dependencies if one camera trap is within one territory while the other is just outside, assuming that territories do not share borders.


## Planning setup

### Planning for power

Starting from a specific **reference or target value for the probability of presence** (i.e., some sort of critical value from an ecological perspective; also the null hypothesis), we can investigate how the **minimal deviation detectable with a desired statistical power** changes with the sample size.
Moreover, we can specify a **critical minimal deviation**, we want to be able to detect and get the required sample size.

We investigate this in the context of the **Binomial test**.
The Binomial test can be used to determine whether a proportion of successes (sightings) in a number of Bernoulli trials (observed sampling locations) differs significantly from an expected proportion of successes (target value). 
The test statistic is the Binomial distribution (@eq-binom).

The following figure visualizes for 100 trials (sampling locations) how much statistical power we have to detect deviations of two different sizes (h_a1 and h_a2) from a certain target (h_0).


```{r fig-plot-binomial-power}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Power of the binomial test (I = 100)"
#| out-width: 150%
#| fig-height: 6
#| fig-width: 10

source("../_functions/plot_binom_power.R")
plot_binom_power(
    n = 100, # number of trials
    psi_0 = 0.05, # probability of succes under h_0
    psi_a = c(0.1, 0.15), # probability of succes under h_a, or vector if multiple values
    alpha = 0.1 # type 1 error
)$plot_power
```


We can look at a whole range of sampling locations and each time determine the deviation that is detected with exactly the desired power.
This is the minimal detectable deviation given the desired power.

For the Chinese muntjac we have the following **specific implementation**:

- target probability of presence: 0, as eradication is the goal
- critical minimal deviation: small, e.g., a probability of presence of 0.05 or 0.01; these mean that animals are expected to be present at 5 or 1 % of all possible sampling locations
- desired statistical power (i.e., the probability of detecting the critical minimal deviation): high, e.g., 0.90
- We rely on the one-sided Binomial test, as the probability of presence cannot be smaller than 0; we set the type-1 error as $\alpha = 0.1$

The following figure shows the minimal detectable probability of presence (in %) as a function of the number of sampling locations.^[
Here we happened to have chosen values for $\alpha$ and power such that $(1- \alpha) = power$ (which of course needs not be the case). 
By coincidence, we discovered that, for this case, the upper bound of the $(1- \alpha)$ Clopper-Pearson confidence interval (CI) for $k = 0$ seemed to correspond to the minimal detectable presence with $(1- \beta)$ power.
Further evidence for this correspondence has been collected via extensive numerical tests.
In addition, there exists a closed form expression for the Clopper-Pearson CI if $k = 0$ for the two-sided case, which can be adapted for the one-sided case [@thulin_cost_2014].
Based on this expressions it can be shown analytically, that indeed $\alpha = (1- power) = \beta$, for a probability of success that equals the upper bound of the $(1- \alpha)$ Clopper-Pearson CI for $k = 0$ (see appendix A).
Using this closed-form expression, we can also solve for the required number of locations given a certain minimal detectable presence.
We use the corresponding solution in producing @fig-minimal-deviation.
Note that the minimal detectable deviation is not generally linked to CI bounds. 
]



```{r fig-minimal-deviation}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Minimal detectable presence as a function of sample size"
#| out-width: 150%
#| fig-width: 5
#| fig-height: 5

# since, by coincidence, alpha = beta = 1 - power, and h_0: psi = 0, 
# we can use the closed-form expression of the one-sided Clopper-Pearson CI
# for k = 0 to get the minimal detectable presence
# otherwise we can use: source("../../_functions/funs_binom_mdd.R")
source("../_functions/funs_binom_ci.R")

# simulate curve
data_mdd <- data.frame(
  nl = seq(1,600),
  power = 0.9
) |>
  dplyr::mutate(
    mdd = purrr::map2_dbl(nl, power, \(x, y){
      get_ciu_k0(nl = x, conf_level = y, one_sided = TRUE) 
    })
  )

# points to highlight
data_mdd_hl <- data.frame(
    mdd = c(0.5, 0.1, 0.05, 0.01, 0.005),
    power = 0.9
  ) |> 
    dplyr::mutate(
      nl = purrr::map2_int(mdd, power, \(x, y){
        get_nl(ciu = x, conf_level = y, one_sided = TRUE) 
        }),
      label = sprintf("%.*f%% presence\n%.*f locations", 
                      ifelse(.data$mdd > 0.005, 0, 1), mdd*100, 0, nl),
      col = ifelse(mdd %in% c(0.01), TRUE, FALSE)
    ) 

# visualize minimal detectable deviation as a function of sample size
plot_mdd <- ggplot2::ggplot(data = data_mdd, ggplot2::aes(x = nl, y = mdd)) + 
  ggplot2::geom_line(linewidth = 1) + 
  ggplot2::geom_point(data = data_mdd_hl, size = 3) + 
  ggforce::geom_mark_circle(
    data = data_mdd_hl, 
    ggplot2::aes(
      x0 = nl + 50, y0 = mdd + 0.1, 
      label = label, group = nl, 
      fill = col
      ) #,
  #  con.type = "straight",
  #  label.fontface = "plain",
  #  label.fontsize = 12,
  #  label.buffer = grid::unit(20, "mm")
    ) +
  ggplot2::scale_y_continuous(labels = scales::percent) + 
  ggplot2::scale_fill_manual(values = c("transparent", "#C27900"), guide = "none") + 
  ggplot2::coord_cartesian(xlim = c(0, 500), ylim = c(0,0.25)) + 
  ggplot2::labs(
    x = "Number of sampling locations",
    y = "Minimal detectable presence (90 % statistical power)"
    ) + 
  ggplot2::theme_bw() + 
  ggplot2::theme(ggplot2::element_text(size = 16)) 

plot_mdd

```

A few possibly interesting probabilities of presence are highlighted: 
For example, if the species is present at 1% of locations then we can detect this with a 90% probability using a sample of `r data_mdd_hl |> dplyr::filter(grepl("1%", label)) |> dplyr::pull(nl)` locations.

### Planning for precision

Alternatively, we could argue that we simply consider the case in which we observe $0$ animals (and what that tells us):^[
Previously, we did not a-priori determine what we would observe.
]
If at least one observation of an animal is made, action has to be taken anyway.
<!-- We now want to know how many zero-observations we need to being able rule out a critical probability of presence.-->
If zero-observations are made across locations, however, we can guess that there are no animals
(i.e., the (maximum likelihood) point estimate of the probability of presence is $0$).
The precision associated with this guess however depends on the number of sampling locations:
we will trust our guess more having 100 as compared to 10 locations with zero-observations.
The more locations without sightings, the less plausible probabilities of presence estimates (much) larger than $0$ become.
We can formalize this by looking at the confidence interval (CI) estimate, which narrows for an increasing numbers of zero-sightings.
Values within the CI are alternative probability of presence values which are still plausible given the respective number of zero-sightings.^[Interpreting single CI estimates is tricky, and I need to double check whether this interpretation is admissible (enough); for now based on Lakens: https://lakens.github.io/statistical_inferences/07-CI.html]
We thus desire the sample size for which the interval does no longer include a critical probability of presence (see previous section).
@fig-minimal-deviation can also be read as displaying the upper bound of the Clopper-Pearson CI for varying numbers of locations with zero-sightings.

Note that this way of approaching the sample size calculation is conceptually not so different from the first approach:
There exists an inverse relationship between a CI and a statistical hypothesis test (ref: Casella & Berger), which also holds for the 
Clopper-Pearson CI and the Binomial test [@thulin_cost_2014; @tony_cai_one-sided_2005]:
we would have retained the null hypothesis had we observed p-values associated with the probability of presence-values contained in the CI and rejected the null hypothesis had we observed p-values associated with the probability of presence-values outside the CI.^[For the specific null hypothesis of zero-probability of presence this seems a bit counter-intuitive: With the Binomial test, we would reject this null hypothesis already with a single sighting (independent of the overall number of trials), while the Clopper-Pearson CI includes values that correspond to non-zero-sightings.
]
The CI bounds thus mark *some* minimal detectable deviation from our best guess corresponding to the earlier null hypothesis or target value.
We can, however, not generally link this to a probability statement, as there is no general link between power and the bounds of CIs.
<!-- Being able to reject a null hypothesis per se does not generally guarantee that we end up with the desired power. -->
This is also why planning for power seems to be more informative.
It is also more intuitive for monitoring scopes aiming at detection, as power is the sensitivity of a design to detect an effect.


Note that in neither of the two approaches can we make statements regarding the probability of the null- or alternative hypothesis being true.
Nor can we interpret the CI in the sense that the true probability of presence lies within the interval with a high probability.
This would require a Bayesian perspective, which is discussed in section XX.
<!-- HERE -->

<!-- For > 58 locations, the CI estimate excludes a probability of presence of 5%, which might be a critical threshold from an ecological perspective. 
A probability of presence of 1% is however still included in the interval.
To exclude that value, we would need to measure at > 298 locations.-->


### Complement: Classification of observable effects

We can further complement previous results - most sensibly the power-based approach in which no specific observed outcome was assumed - by taking estimation precision of all possible observable effects into account:
For each sample size, we can for all possible sample configurations (i.e., all possible numbers of sightings^[
If we sample 2 locations, we can observe the species at 0, 1, or 2 locations, corresponding to observable probabilities of presence of 0, 0.5 and 1
]) obtain the observable effects (observable probabilities of presence) and their CI estimates.
Based on width and position of the CIs with respect to the target and the minimal detectable deviation, we can categorize each observable effect according to the [effectclass-logic](https://inbo.github.io/effectclass/articles/visualisation.html#format_ci-and-classification).
The following figure illustrates this and shows which interpretations we can afford given different sample sizes and observed probabilities of success.
Each point corresponds to an observable effect, its colour codes its CI-derived classification.
The minimal detectable presences are marked via a black line.

Note that this is not about but can complement sample-size calculations.


<!--- fix threshold issue ---->
<!--- use points ---->


```{r fig-classification}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Classification of observable effects"
#| out-width: 150%
#| fig-height: 5
#| fig-width: 10

source("../sample_design/plot_effectclass.R")

plot_ec

```


