## Sample size planning for detection {#sec-samplesize-detection}

### Data-generating model

For the monitoring needs "detect spread under management" and "detect presence after management", we stick to certain **data-generating distributions** as the data-generating models.
In @sec-samplesize-decline, we combine data-generating distributions with regression modelling to further decompose the parameters of the data-generating distribution.

#### Data-generating distribution

##### Bernoulli and binomial distribution {#sec-bernbinom}

The monitoring needs related to detection concern the presence/absence of animals as outcomes. 
We can rely on the **Bernoulli distribution** as the data-generating distribution. 
The parameter of focus is the probability of success, here the probability of presence:

$$
obs_{i,t=t'} \sim Bern(\psi_{i,t=t'})
$$ {#eq-bern1}

- $obs_{i,t=t'}  \in \{0,1\}$: observation of animals at location $i$, year $t'$, aggregated over days
- $\psi_{i,t=t'}  \in [0,1]$: probability of presence at location $i$, year $t'$
  - matching above aggregation (i.e., it concerns the presence of at least one animal on at least ...) 

We can think of the probability of presence featuring in @eq-bern1 as being an observed one, which should be linked to a true probability of presence.
Ideally the link is a one-to-one link.
An imperfect detection probability of the monitoring method can weaken the link. 
For now we assume a camera trap detection probability of one, hence we can perfectly detect animals if they are present.
<!--The same animals being present multiple times or multiple animals being present is accommodated by the above 'aggregated interpretation' of the probability of presence.
The same animals being present multiple times or multiple animals being present should not be an issue and can be accommodated by thinking of $\psi$ as the probability of at least one animal being present.-->

If we assume that the probability of presence is **invariant across locations** (i.e., $\psi_{i,t=t'} = \psi_{t=t'}$),
it follows from above that the number of locations with animal sightings (i.e., $nobs_{t=t'} := \sum_{i=1}^I obs_{i,t=t'}$) can be thought of as realisations of the **binomial distribution** in function of $\psi_{t=t'}$:

$$
nobs_{t=t'} \sim Binom(\psi_{t=t'}, I)
$$ {#eq-binom}

- $nobs_{t=t'}  \in \{0,1,2,3, ..., I\}$: number of observations of animals summed across all locations, at year $t'$
- $\psi_{t=t'}  \in [0,1]$: probability of presence at time $t = t'$
  - here assumed invariant across all locations
- $I$: the total number of locations

If we allow for **location-specific probabilities of presence**, the situation gets slightly more complicated.
In that case $nobs_{t=t'}$ could have been generated by the **Poisson binomial distribution**, which captures a set of Bernoulli trials with unequal probabilities of success/presence.
There is no particular assumption made about how the different probabilities of presence arise (i.e., they are assumed to be fixed in statistical terms).
Importantly, the distribution can be approximated by a binomial distribution, with a probability of presence corresponding to the mean probability of presence across locations (see @fig-plot-poisbinom for an example; the error of approximation decreases with a decreasing variance of the location-specific probabilities of presence).
This means that the following results, which hold exactly for the location-unspecific case, hold approximately for the location-specific case.
<!-- Beta binomial distribution: https://stats.stackexchange.com/questions/390023/poisson-binomial-vs-beta-binomial]-->


```{r fig-plot-poisbinom}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Estimated densities of the Poisson binomial (solid) vs. approximating binomial distribution (dashed)"
#| fig-height: 4.5

# function to draws from the poisson binomial distribution
source("../_functions/rpoisbinom.R")

# number of locations
I = 100

data_poisbin <- lapply(
  seq(1,5000),
  \(i){
    rpoisbinom(
      I = I,
      p_min = 0, p_max = 0.5,
      runif_seed = 123, rbinom_seed = i
      ) |>
      cbind(i = i)
  }
) |> dplyr::bind_rows()

# density poisson binomial
tmp <- data_poisbin$y_sum |> density()
dens_poisbin <- data.frame(
  x = tmp$x,
  y = tmp$y
)

# density binomial
dens_binom <- lapply(
  seq(0,I),
  \(i){
    data.frame(
      x = i,
      y = dbinom(x = i, size = I, prob = data_poisbin$psi_mean[1])
    )
  }
) |> dplyr::bind_rows()

ggplot2::ggplot(data = dens_binom, ggplot2::aes(x = x, y = y)) + 
  ggplot2::geom_line(linetype = "dashed") + 
  ggplot2::geom_line(data = dens_poisbin) + 
  ggplot2::theme_bw() + 
  ggplot2::labs(
    x = "Number of successes",
    y = "Probability"
  )

```

Finally, one assumption should be highlighted:
Under the binomial distribution there are **no dependencies between locations**.
This is a defensible assumption if camera traps are located relatively far apart, especially if a spatially balanced sample is taken.
The assumption becomes more questionable if the sample size increases such that camera traps stand closer together:
in that case, we might expect positive dependencies if two camera traps are within one territory of one individual, 
and negative dependencies if one camera trap is within one territory while the other is just outside, assuming that territories do not share borders.
Given the sizes of the sample frames of the distribution and buffer area (@sec-sampling), 
the numbers of sampling locations needs to be rather high for their configuration to become too dense.

### Planning scenarios

#### Planning for power {#sec-planning-power}

Having established the context for sample size planning, we can now proceed with the first planning scenario, planning for power.
More specifically, starting from a specific **reference or target value for the probability of presence** (i.e., some sort of critical value from an ecological perspective; also the null hypothesis), we can investigate how the **minimal deviation detectable with a desired statistical power** changes with the **sample size**.
Moreover, we can specify a **critical minimal deviation**, we want to be able to detect and get the **required sample size**.

We investigate this in the context of the **binomial test**.
The binomial test can be used to determine whether a proportion of successes (here: animal sightings) in a number of Bernoulli trials (here: observed sampling locations) differs significantly from an expected proportion of successes (here: the chosen target value). 
The test statistic is the above-discussed binomial distribution (@eq-binom).

The following figure visualizes for 100 trials (sampling locations) how much power we have to detect deviations of two different sizes (h_a1 and h_a2) from a certain target (h_0).


```{r fig-plot-binomial-power}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Power of the binomial test (I = 100)"
#| fig-height: 8

source("../_functions/plot_binom_power.R")
plot_binom_power(
    n = 100, # number of trials
    psi_0 = 0.05, # probability of success under h_0
    psi_a = c(0.1, 0.15), # probability of success under h_a, or vector if multiple values
    alpha = 0.1 # type 1 error
)$plot_power + 
  ggplot2::theme(legend.position = "bottom")
```


We can now look at a whole range of sampling locations and each time determine the deviation that is detected with exactly the desired power.
This is the minimal detectable deviation given the desired power.

For the Chinese muntjac we have the following **specific implementation**:

- target probability of presence: 0, as eradication is the goal
- critical minimal deviation: small, e.g., a probability of presence of 0.05 or 0.01; these mean that animals are expected to be present at 5 or 1 % of all possible sampling locations
- desired power (i.e., the probability of detecting the critical minimal deviation): high, e.g., 0.90
- We rely on the one-sided binomial test, as the probability of presence cannot be smaller than 0; we set the type-1 error as $\alpha = 0.1$

The following figure shows the minimal detectable probability of presence (in %) as a function of the number of sampling locations.^[
Here we happened to have chosen values for $\alpha$ and power such that $(1- \alpha) = power$ (which of course needs not be the case). 
By coincidence, we discovered that, for this case, the upper bound of the $(1- \alpha)$ Clopper-Pearson confidence interval (CI) for $k = 0$ seemed to correspond to the minimal detectable presence with $(1- \beta)$ power.
Further evidence for this correspondence has been collected via extensive numerical tests.
In addition, there exists a closed form expression for the Clopper-Pearson CI if $k = 0$ for the two-sided case, which can be adapted for the one-sided case [@thulin_cost_2014].
Based on this expressions it can be shown analytically, that indeed $\alpha = (1- power) = \beta$, for a probability of success that equals the upper bound of the $(1- \alpha)$ Clopper-Pearson CI for $k = 0$ (see appendix A).
Using this closed-form expression, we can also solve for the required number of locations given a certain minimal detectable presence.
We use the corresponding solution in producing @fig-minimal-deviation.
Note that the minimal detectable deviation is not generally linked to CI bounds. 
]



```{r fig-minimal-deviation}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Minimal detectable presence as a function of sample size"
#| fig-asp: 0.618

# since, by coincidence, alpha = beta = 1 - power, and h_0: psi = 0, 
# we can use the closed-form expression of the one-sided Clopper-Pearson CI
# for k = 0 to get the minimal detectable presence
# otherwise we can use: source("../../_functions/funs_binom_mdd.R")
source("../_functions/funs_binom_ci.R")

# simulate curve
data_mdd <- data.frame(
  nl = seq(1,600),
  power = 0.9
) |>
  dplyr::mutate(
    mdd = purrr::map2_dbl(nl, power, \(x, y){
      get_ciu_k0(nl = x, conf_level = y, one_sided = TRUE) 
    })
  )

# points to highlight
data_mdd_hl <- data.frame(
    mdd = c(0.5, 0.1, 0.05, 0.01, 0.005),
    power = 0.9
  ) |> 
    dplyr::mutate(
      nl = purrr::map2_int(mdd, power, \(x, y){
        get_nl(ciu = x, conf_level = y, one_sided = TRUE) 
        }),
      label = sprintf("%.*f%% presence\n%.*f locations", 
                      ifelse(.data$mdd > 0.005, 0, 1), mdd*100, 0, nl),
      col = ifelse(mdd %in% c(0.01), TRUE, FALSE)
    ) 

# visualize minimal detectable deviation as a function of sample size
plot_mdd <- ggplot2::ggplot(data = data_mdd, ggplot2::aes(x = nl, y = mdd)) + 
  ggplot2::geom_line(linewidth = 1) + 
  ggplot2::geom_point(data = data_mdd_hl, size = 3) + 
  ggforce::geom_mark_circle(
    data = data_mdd_hl, 
    ggplot2::aes(
      x0 = nl + 50, y0 = mdd + 0.1, 
      label = label, group = nl, 
      fill = col
      ) #,
  #  con.type = "straight",
  #  label.fontface = "plain",
  #  label.fontsize = 12,
  #  label.buffer = grid::unit(20, "mm")
    ) +
  ggplot2::scale_y_continuous(labels = scales::percent) + 
  ggplot2::scale_fill_manual(values = c("transparent", "#C27900"), guide = "none") + 
  ggplot2::coord_cartesian(xlim = c(0, 500), ylim = c(0,0.25)) + 
  ggplot2::labs(
    x = "Number of sampling locations",
    y = "Minimal detectable presence (90 % statistical power)"
    ) + 
  ggplot2::theme_bw() + 
  ggplot2::theme(ggplot2::element_text(size = 16)) 

plot_mdd

```

A few possibly interesting probabilities of presence are highlighted: 
For example, if the species is present at 1% of locations then we can detect this with a 90% probability using a sample of `r data_mdd_hl |> dplyr::filter(grepl("1%", label)) |> dplyr::pull(nl)` locations.

#### Planning for precision (variant 1)


Alternatively, we could argue that we simply consider the case in which we observe $0$ animals and what we can infer in that case^[
Previously, we did not a-priori determine what we would observe.
]:
if at least one observation of an animal is made, action has to be taken anyway.
<!-- We now want to know how many zero-observations we need to being able rule out a critical probability of presence.-->
If $i'$ zero-observations are made across locations, however, we can guess that there are no animals
(i.e., the (maximum likelihood) point estimate of the probability of presence is $0$).
The precision associated with this guess however depends on the number of sampling locations.
We will trust our guess more having 100 as compared to 10 locations with zero-observations.
The more locations without sightings, the less plausible probabilities of presence estimates (much) larger than $0$ become.
We can formalize this by looking at the confidence interval (CI) estimate, which narrows for an increasing numbers of zero-sightings.
Values within the CI are alternative probability of presence values which are still plausible given the respective number of zero-sightings.^[Interpreting single CI estimates can be tricky. The present attempt is based on [this blogpost](https://lakens.github.io/statistical_inferences/07-CI.html).]
We thus desire the sample size for which the interval does no longer include a critical probability of presence (see previous section).
Given the here employed settings, @fig-minimal-deviation can also be read as displaying the upper bound of the Clopper-Pearson CI for varying numbers of locations with zero-sightings (@sec-planning-power).

As a side note: leaving the specific settings chosen aside, this planning scenario is conceptually not so different from the scenario planning for power, in the following sense.
There exists an inverse relationship between a CI and a statistical hypothesis test [@casella_statistical_2002], which should also hold for the 
Clopper-Pearson CI and the binomial test [@thulin_cost_2014; @tony_cai_one-sided_2005]:
we would have retained the null hypothesis had we observed p-values associated with the probability of presence-values contained in the CI and rejected the null hypothesis had we observed p-values associated with the probability of presence-values outside the CI.^[For the specific null hypothesis of zero-probability of presence this seems a bit counter-intuitive: With the binomial test, we would reject this null hypothesis already with a single sighting (independent of the overall number of trials), while the Clopper-Pearson CI includes values that correspond to non-zero-sightings. To be investigated further in the future.
]
The CI bounds thus mark *some* minimal detectable deviation from our best guess corresponding to the earlier null hypothesis or target value.
We can, however, not generally link this to a probability statement, as there is no *general* link between power and the CI bounds.
<!-- Being able to reject a null hypothesis per se does not generally guarantee that we end up with the desired power. -->
This is also why planning for power seems to be more informative.
It is also more intuitive for monitoring aiming at detection, as power is the sensitivity of a design to detect an effect.


Finally, note that in neither of the two planning scenarios can we make statements regarding the probability of the null- or alternative hypothesis being true.
Nor can we interpret the CI in the sense that the true probability of presence lies within the interval with a high probability.
This would require a Bayesian perspective, which is discussed in @sec-samplesize-eradication.


#### Planning for precision (variant 2) 

Another variant of precision-based planning that might mainly be used to complement results from the power-based scenario 
takes into account **estimation precision of observable effects**:
For each sample size, we can for all possible sample configurations (i.e., all possible numbers of sightings^[
If we sample 2 locations, we can observe the species at 0, 1, or 2 locations, corresponding to observable probabilities of presence of 0, 0.5 and 1
]) obtain the observable effects, that is, observable probabilities of presence, and their CI estimates.
This is done by simulating data under all sample configurations and sizes and calculating the relative frequencies of presences per sample to obtain the observable probabilities of presence.
The CI estimates are then obtained by subjecting each sample to the binomial test, which, at least if the base-R function *binom.test()* is used, also returns estimates of the Clopper-Pearson CI.
Based on width and position of the CI bounds with respect to the target value and the minimal detectable deviation, we can categorize each observable effect according to the [effectclass-logic](https://inbo.github.io/effectclass/articles/visualisation.html#format_ci-and-classification):
if the lower bound of the CI is not smaller than the target value while its upper bound is smaller than the minimal detectable deviation, an observable effect can be classified as "potentially over target";
if the lower bound of the CI is larger than the target value while its upper bound is also larger than the minimal detectable deviation, an observable effect can be classified as "over target";
if both the lower and the upper CI bound exceed the minimal detectable deviation, an observable effect can be classified as "strongly over target".

@fig-classification illustrates this approach and thus shows which interpretations we can afford given different sample sizes and observed probabilities of success.
Each point corresponds to an observable effect, its colour codes its CI-derived classification.
The minimal detectable presences obtained under the power-based scenario are marked via a black line.
The figure is complete for the displayed sample sizes in the sense that it contains all possible observable effects. 


```{r fig-classification}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Classification of observable effects"
#| fig-asp: 0.618

source("../sample_design/plot_effectclass.R")
plot_ec

```


Results obtained under this variant of precision-based planning can be used if critical minimal deviations (@sec-planning-power) are desired that *additionally* classify as "strongly over target".
As @fig-classification shows, the corresponding sample sizes will be larger than the ones obtained under the planning for power scenario (contrasting the black line with the borderline between effects "strongly over target" vs. "over target").

