# Sample size planning - scope A {#sec-scope-a}

## Data-generating model

### Data-generating distribution

#### Poisson distribution

Monitoring scope A (confirm population decline under management) most naturally concerns abundances of animals as outcomes (i.e., numbers of individuals per location), so we can rely on the **Poisson distribution**, with the parameter of focus being the expected number of observed animals:

$$
nobs_{i,t} \sim Pois(\mu_{i,t})
$$ {#eq-pois}
with

- $nobs_{i,t} \in \{0,1,2,3, ...\}$: number of observations of animals at location $i$, year $t$
  - aggregated over days by using e.g. de maximum number of animals on a day
- $\mu_{i,t} \in (0, \infty)$: expected number of observations of animals at location $i$, year $t$ 

We assume that $\mu_{i,t}$ is linked to the expected true population size / the true abundance at location $i$, year $t$, $\nu_{i,t}$.^[Following the logic of N-mixture modeling, or latent variable modeling, one would rather assume that $nobs_{i,t}$ is linked to the true population size at location $i$, year $t$, $n_{i,t}$; if the link holds at that level then it should also be implied at the level of expectations, but not the reverse.]
For simplicity, we assume a linear relationship that is invariant across all dimensions considered (i.e., space and time), but also across abundance itself:

$$
\mu_{i,t} = a * \nu_{i,t}
$$ {#eq-pois2}

In an ideal case $a = 1$, which means we can make statements about absolute abundances over space, time and abundance.
If $a \neq 1$, then we can still make statements about relative abundances over space, time and abundance *without having to know* $a$.
Reasons for $a \neq 1$ can be: an imperfect detection probability; animals being recorded more than once within a day. 

However, the assumption of $\mu_{i,t}$ being linked to $n_{i,t}$ might be problematic if animals are very territorial, with no or little overlap between territories.
Then we are most likely only observing one to a few individuals and the number of observations is more related to their activity during that day.
If abundances are basically one at this geographical scale, then we are dealing with data of presence and absence.

<!--HERE: Here we have 2 problems: abundances rather 1; multiple observations more likely to pertain to the same animal(s) than to different ones.-->

#### Bernoulli distribution

If abundance at the given geographical scale is rather about the presence of animals, then we can alternatively rely on the **Bernoulli distribution** as the data-generating distribution (cf. @sec-bernbinom and @eq-bern1):

$$
obs_{i,t} \sim Bern(\psi_{i,t})
$$ {#eq-bern2}

- $obs_{i,t}  \in \{0,1\}$: observation of animals at location $i$, year $t$, aggregated over days
- $\psi_{i,t}  \in [0,1]$: probability of presence at location $i$, year $t$


### Regression model

#### Logistic regression

Starting from the distributional assumption in @eq-bern2, we can use the **Logistic regression model** 
to linearly decompose $\psi_{i,t}$, the parameter of the Bernoulli distribution, after applying the logit-link function:


$$
logit(\psi_{i,t}) = log(\frac{\psi_{i,t}}{1-\psi_{i,t}}) = \eta_{i,t} \\
$$ {#eq-logreg1}
with:

- $logit(\cdot)$: logit-link function (i.e., the log-odds/ logarithm of the odds)
- $log(\cdot)$: natural logarithm
- $\frac{\psi_{i,t}}{1-\psi_{i,t}}$: odds of presence
- $\eta_{i,t}$: linear predictor

The **most simple regression model** variant includes an intercept and a linear effect of year, both location unspecific, so index $i$ is dropped:

$$
\eta_{i,t} = \beta_{0} + \beta_{1} * year_t^*
$$ {#eq-logreg2}
with:

- $\beta_{0}$: intercept; linear predictor if $year_t^*= 0$
- $\beta_{1}$: slope associated with $year_t^*$
- $year_t^*$: some linear function of year

The **parameter of interest** - corresponding to our monitoring target 'trend' - is obviously the **slope**, $\beta_{1}$.
It can be interpreted in terms of the linear predictor,
in terms of the odds of presence ($odds_{i,t}: exp(\eta_{i,t}) = \frac{\psi_{i,t}}{1-\psi_{i,t}}$),
or in terms of the probability of presence ($\psi_{i,t} = \frac{odds_{i,t}}{1+odds_{i,t}}$).

The interpretation in terms of the linear predictor is the one we are familiar with from simple linear regression:
$\beta_1$ reflects the absolute change in linear predictor values given a one-unit change in year*.

The interpretation in terms of the odds of presence needs to incorporate the natural logarithm linking the odds of presence to the linear predictor:
Here, $exp(\beta_1)$ is the factor by which the odds of presence change given a one-unit change in year (see Appendix B).
Also, $(exp(\beta_1)-1)*100$ is the percentage change in the odds.

The interpretation in terms of the probability of presence...<!-- HERE -->


The following figure shows a number of different trends implied by different sizes of $\beta_1$ for the linear predictor, $\eta$, the odds and the probability of presence, $\psi$.
What a critical effect size is needs to be determined from a biological perspective.

```{r fig-size-trend-2}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Visualization of different effect sizes of beta_1 in the logistic regression model"
#| out-width: 150%

data_trend <- data.frame(
  # size of linear effect of year
  effect_size = -seq(0, 2, 0.2)
) |>
  tidyr::crossing(
   year_star = seq(0,15)
) |>
  dplyr::mutate(
    eta =  effect_size*year_star + 2,
    odds = exp(eta),
    psi = odds / (1 + odds),
    effect_size = factor(effect_size) |> forcats::fct_rev()
    ) |>
  tidyr::pivot_longer(
    cols = tidyselect::all_of(c("eta", "odds", "psi")),
    names_to = "out_name",
    values_to = "out_value"
  )
ggplot2::ggplot(
  data = data_trend, 
  mapping = ggplot2::aes(x = year_star, y = out_value, group = effect_size)
  ) + 
  ggplot2::geom_line(ggplot2::aes(color = effect_size), linewidth = 1.2) +
  ggplot2::facet_wrap(~ out_name, scales = "free_y", ncol = 3) + 
  ggplot2::labs(x = "year*", y = "outcome", color = "size beta1") +
  ggplot2::scale_x_continuous(breaks = data_trend$year_star |> unique()) + 
  ggplot2::scale_colour_viridis_d(direction = -1) + 
  ggplot2::theme_bw()

```


If $\beta_1 = -0.2$, then the odds change by a factor of `r round(exp(-0.2), digits = 2)` per year, 
and if $\beta_1 = -1$, that factor is `r round(exp(-1), digits = 2)`. 
<!-- HERE: add interpretation in percent (see Poisson) -->

Finally, relaxing the assumption of a location-unspecific regression model, we can extend the regression model @eq-logreg2 by **random effects across locations**:

At level 1 (within locations) we have:

$$
\eta_{i,t} = \beta_{0,i} + \beta_{1,i} * year_t^*
$$ {#eq-logreg3}
with:

- $\beta_{0,i}$: random intercept, varying across locations <!-- location-specific intercept-->
- $\beta_{1,i}$: random slope associated with $year_t^*$, varying across locations


At level 2 (between locations) we have:

$$
\beta_{0,i} = \gamma_{00} + u_{0i}\\
\beta_{1,i} = \gamma_{10} + u_{1i}\\ 
$$
with:

- $\gamma_{00}$: fixed part of intercept; mean of intercepts across locations; 
- $u_{0i}$: random part of intercept; locations-specific deviations from fixed intercept
- $\gamma_{10}$: fixed part of slope
- $u_{1i}$: random part of slope; locations-specific deviations from fixed slope

The level-2 errors $u_{0i}$ and $u_{1i}$ are assumed to be jointy normally distributed with zero mean vector and valid covariance matrix.

## Planning setup

### Planning for power

Based on the regression models in @eq-logreg2 and @eq-logreg3, we can - given a zero-slope as the reference - determine the minimal slope we can detect with a high statistical power across a range of interesting / relevant number of locations.

We address this via a **simulation-based power analysis**, using the R package *mlpwr* (ref.).
The package provides support for simulation-based power analyses, with the capacity of simultaneously tackling multiple design dimensions (such as the number of sampling units at different levels, the number of time points in a longitudinal study, the spacing in time of repeated measures etc.) as well as budget constraints.
It can be used to identify a power-optimal design, 
a power-optimal design among the designs with the desired cost, 
or a cost-optimal design among the designs with the desired statistical power.
It combines regular Monte Carlo simulations with a model-driven search of the design space.^[
The proposed algorithm works as follows:
In the start-up phase, power is simulated for a number of systematically selected designs.
This is followed by a prediction and updating phase, which are looped until a certain stopping criterion is reached.
In the prediction phase a model is estimated based on the initially simulated power data to approximate the general relationship between the design dimensions and power.
The estimated model covers the entire parameter space and can then be used to search for a (conditionally) optimal solution
This is done via numerical optimisation based on a loss function for either power or costs.
In the updating phase newly simulated data are added at the earlier determined optimal location.
The two phases are iteratively repeated until either the maximum number of iterations is reached, or the model predictions are sufficiently accurate.
]
The authors refer to their approach as 'surrogate modeling' because the model predictions for power are used as a surrogate for those designs that have not (yet) been simulated.
The package features different surrogate models depending on the number of design parameters and the user's choice.
A search model-based strategy is more efficient than other search methods (e.g., a grid search), 
particularly if one deals with multiple design dimensions.
<!-- HERE: add that other optimization criteria are possible too, e.g. precision -->

Here we slightly re-purpose the package:
Instead of optimising the number of sampling locations for statistical power, given a certain type-1 error rate and certain effect sizes, we **optimise the parameter of interest, the slope, for power**, given a certain type-1 error rate, certain effect sizes (of the remaining parameters) and a certain sample size.
This should work as the two tasks are not so different:
Instead of looking for the minimal sample size that meets the set criteria,
we look for the minimal effect size that meets the set criteria.
Also, the relationship between the effect size and power should be of similar nature as the relationship between sample size and power 
(e.g., power as a smooth, monotonically increasing function of increasing sample or effect size).^[
In appendix X, we compare the exact results obtained for the Binomial test with results obtained via mlpwr.
]

```{r get-mlpwr-res}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'

# load results
load("../../output/sample_design/mlpwr_logreg/gridpars_20251022-133925.Rda")
load("../../output/sample_design/mlpwr_logreg/res_list_20251022-133925.Rda")
  
```


For our analysis, we use the following **settings for the design and model parameters** (@eq-logreg3):

- sample sizes are set to: `r grid_pars$n |> unique() |> paste0(collapse = ", ")`
- the numbers of years across which trends are estimated are set to: `r grid_pars$x_end |> unique() |> paste0(collapse = ", ")`
  - hence, we for instance have $year_t^* \in \{`r seq(0, grid_pars$x_end[1]) |> paste0(collapse = ", ")`\}$
- the fixed intercept, $\gamma_{00}$, is set to: `r grid_pars$b0 |> unique() |> paste0(collapse = ", ")`
  - this corresponds to initial probabilities of the species being present of: `r grid_pars$b0 |> unique() |> purrr::map_vec(.x = _, \(x){exp(x)/(1+exp(x))} |> round(2)) |> paste0(collapse = ", ")`
- the random intercept variance is set to: `r grid_pars$u0_var |> unique() |> paste0(collapse = ", ")`
- the random slope variance is set to: `r grid_pars$u1_var |> unique() |> paste0(collapse = ", ")`
- the random intercept-slope covariance is set to: `r grid_pars$u0u1_covar |> unique() |> paste0(collapse = ", ")`

These settings are fully crossed, resulting in `r grid_pars |> nrow()` conditions.
Across all conditions, statistical power is held at $\pi =`r grid_pars$power |> unique()`$,
and the type-1 error at $\alpha = `r grid_pars$alpha |> unique()`$.

Concerning the settings of the *mlpwr::find.design()* function we stick to the default settings (e.g., the logistic regression model as the surrogate model) with two important exceptions:

- argument 'integer' is set to 'FALSE' as we are optimizing the fixed slope size, and thus a real-valued "design" parameter
- the lower bounds for the design space (cf. argument 'boundaries') are set to reasonable values in function of the sample size and size of the fixed intercept (initial analysis results revealed malperformance of the surrogate model if lower bounds are too large as the relevant variance in the predictor is then limited to a very narrow value range; lower bounds should however not be too small as then the optimal solution might not be included in the design space.)

Finally, if the random intercept variance is non-zero, then a random intercept fixed slope logistic regression model is estimated, 
otherwise a fixed effects logistic regression model is estimated.


```{r fig-mlpwr}
#| include: true
#| cache: true
#| message: false
#| warning: false
#| results: 'hide'
#| fig-cap: "Results of the mlpwr-power analysis"
#| out-width: 150%
#| fig-height: 10
#| fig-width: 8

library(patchwork)
source("../sample_design/power_mlpwr_logreg_vis.R")

(plot_res/plot_res_odds) + 
  patchwork::plot_annotation(tag_levels = 'A')

```

The **simulation results** are shown in @fig-mlpwr in terms of the minimal detectable slope (panel A) and the minimal detectable percentage change in the odds of presence (panel B).
We observe the following to be expected relationships:

- with more sampling locations, smaller slopes / odds changes can be detected
- with a longer study duration, smaller slopes / odds changes can be detected
- with smaller fixed intercept / baseline odds, slopes / odds changes must be larger to be detected
- no clear effect of the random intercept variance (comparison also concerns different models fitted)

For selected parameter values, @fig-simdata displays example data simulated from the random-intercept logistic regression model for $I = 100$ locations.
This is an illustration of patterns arising under the minimal detectable slope in different scenarios.


```{r fig-simdata}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Data simulated from the random-intercept logistic regression model (I = 100)"
#| out-width: 150%

source("../_functions/sim_logreg.R")

# simulate data
data_y <- res_vis |>
  dplyr::mutate(block = rep(seq(1,8), each = 9), .after = index) |>
  dplyr::filter(!is.na(mdd), x_end == min(x_end), u0_var > 0) |>
  dplyr::filter(n %in% c(min(n), max(n)), .by = block) |>
  dplyr::select(c("n", b0_true = "b0", "mdd", "x_end", "u0_var")) |>
  dplyr::mutate(
    simdata = purrr::pmap(
      list(
        b0 = b0_true,
        b1 = mdd,
        x_end = x_end, # using seq here causes trouble
        n = 100,
        u0_var = u0_var
      ),
      \(b0, b1, x_end, n, u0_var) {
        sim_logreg(b0 = b0, b1 = b1, x = seq(0, x_end), n = n, u0_var = u0_var)
        }
    )
  ) |>
  tidyr::unnest(simdata) 

# visualize
data_y |>
  dplyr::mutate(
    cond_fct = paste0("b0 = ",b0, ",\nn = ", n, ",\nb1 = ", round(b1, 2)),
    cond_fct = factor(cond_fct, unique(cond_fct))
  ) |>
ggplot2::ggplot(
  data = _,
  mapping = ggplot2::aes(x = x, y = y, group = id)
) +
  ggplot2::geom_line(ggplot2::aes(y = psi), linetype = "longdash", linewidth = 0.5, alpha = .5) +
  ggplot2::geom_line(ggplot2::aes(color = id), linewidth = 1, alpha = .2,
                     position = ggplot2::position_jitter(w = 0.03, h = 0.03)) +
  ggplot2::facet_wrap(facets = ggplot2::vars(cond_fct), ncol = 2) +
  ggplot2::labs(x = "year*", y = "outcome", color = "id") +
  ggplot2::scale_x_continuous(breaks = data_y$x |> unique()) +
  ggplot2::scale_colour_viridis_d(direction = -1, guide = "none") +
  ggplot2::theme_bw() + 
  ggplot2::theme(strip.background = ggplot2::element_blank())


```




