---
title: "Sample design for the Chinese muntjac<br>[draft version]"
author: "Janne Adolf"
format: 
  html:
    toc: true
    toc-location: right-body
    toc-depth: 3
    toc-expand: true
    theme: yeti
    number-sections: true
    number-depth: 3
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
lightbox: true
bibliography: ../../../media/MIAS.bib
execute: 
  eval: true
  echo: true

---

```{r setup}

maplist_3857 <- readRDS("../../../data/gis/sample/sampleframe_maplist_3857.Rds")
maplist_31370 <- readRDS("../../../data/gis/sample/sampleframe_maplist_31370.Rds")

mapview::mapviewOptions(basemaps.color.shuffle = FALSE)

```


<!----------------------------------------------------------------------------------------------------------------------------------------------->
<!-- Overview ----------------------------------------------------------------------------------------------------------------------------------->
<!----------------------------------------------------------------------------------------------------------------------------------------------->

# Introduction

## Overview

The present documents evaluates and/or illustrates options for power or precision analysis for the following **monitoring scopes** relevant for the Chinese muntjac:

- A: Abundance under management - trend
  - Estimate changes in population size in relation to sustained eradication measures in distribution area
  - Can concern changes in absolute or relative abundance
  - to confirm management is correlated with decline in population size
- B: Abundance under management - status
  - Determine population size under sustained eradication measures in distribution area in province of Antwerp
  - to confirm eradication at some point
- C: Detection under management - status
  - Detect potential spread into buffer area around distribution area in relation to sustained eradication measures in distribution area
  - Species quasi absent from buffer area
- D: Detection after management - status
  - Detect potential left-over of re-introduced animals in above distribution area after eradication has been completed


Scopes C and D differ in the targeted area, but are otherwise comparable enough to here subsume them.
We thus start with scope combination C & D, and then move on to scopes A and B.  

Per monitoring scope (combination) the following aspects of a (simulation-based) power analysis are discussed:

<!-- HERE -->
- Data-generation
  - What is an appropriate data-generating (and later analysis) model?
- Analysis
  - Thoughts on possible analysis setups (rationale and implementation)
- Sampling
  - How do (simulated) data map onto geographical locations?
  

## Some technical remarks
  
<!-- HERE -->
We assume we have as a starting point data on observed abundances or presences/absences.
The preprocessing pipeline from camera recordings to these data is not considered in the present document.
<!-- [footnote on how it could be relevant?] -->
The data are assumed to vary along the following dimensions:

- $i = 1, ..., I$ sampling locations
- $d = 1, ..., D$ days per measurement period (more precisely: $d_{t}$; here the same for all $t$, so subscript $t$ is left out)
- $t = 1, ..., T$ years, with one measurement period per year

One aggregation is carried out, along the dimension days.
Here we assume that we aggregate across 30 days (standard).
<!-- HERE -->
<!-- Seems conventional approach (see Bollen).-->

We do not consider the case of multiple measurement periods per year.
In that case, differences between periods within years could be allowed in the model, capturing seasonal deviations from a longer-term (i.e., year-level) trend (e.g., more individuals before than after winter).
For now, we however simply assume that the optimal period per year is used to obtain measurements.
Optimality is for instance tied to a maximal detection probability, the absence of complicating factors such as ongoing control actions during measurement and so on.
Status is then determined per year.
Trend between years.

<!-- HERE -->
<!--
We can think about possible data-generating models in the context of the generalized linear (mixed) modeling framework.
Sticking to the not-too-complex model variants, we can rely on well-known properties/behavior and existing implementations for both data generation and model estimation. 
(Whether the model is feasible to implement as a data-analysis model should be considered already at this stage to not complicate things later.)
We do this for abundances.
For presences, we revert to a simpler option.
For now we generally assume that we use the same model for data generation as we use for data-analysis.
[data-generating distribution + regression model]
-->


<!----------------------------------------------------------------------------------------------------------------------------------------------->
<!-- Scopes C & D ---------------------------------------------------------------------------------------------------------------------------->
<!----------------------------------------------------------------------------------------------------------------------------------------------->


# Scopes C & D - data generation & analysis setup

::: {.callout-note icon=false}

**Summary**

[Summary here]

:::

## Data-generating model

For this scope combination, we stick to certain **data-generating distributions** as the data-generating models.
In later sections, we combine data-generating distributions with regression modelling.^[...to further decompose the parameters of focus of the data-generating distributions.]

Monitoring scopes C and D concern detection and thus the presence/absence of animals as outcomes. 
We can rely on the **Bernoulli distribution** as the data-generating distribution. 
The parameter of focus is the probability of success, here the probability of presence:

$$
obs_{i,t=t'} \sim Bern(\psi_{i,t=t'})
$$ {#eq-bern1}

- $obs_{i,t=t'}  \in \{0,1\}$: observation of animals at location $i$, year $t'$
  - aggregated over days: at least one animal on at least one camera frame on at least one day
- $\psi_{i,t=t'}  \in [0,1]$: probability of presence at location $i$, year $t'$
  - matching above aggregation (i.e., it concerns the presence of at least one animal on at least ...) 

We can here think of the probability of presence featuring in @eq-bern1 as being an observed one, which should be linked to a true probability of presence.
Ideally the link is a one-to-one link.
An imperfect detection probability can weaken the link (for now we assume a detection probability of one, hence we can perfectly detect animals if they are present).
<!--The same animals being present multiple times or multiple animals being present is accommodated by the above 'aggregated interpretation' of the probability of presence.-->

If we assume that the probability of presence is **invariant across locations** (i.e., $\psi_{i,t=t'} = \psi_{t=t'}$),
it follows from above that the number of locations with animal sightings (i.e., $nobs_{t=t'} := \sum_{i=1}^I obs_{i,t=t'}$) can be thought of as realisations of the **Binomial distribution** in function of $\psi_{t=t'}$:

$$
nobs_{t=t'} \sim Binom(\psi_{t=t'}, I)
$$ {#eq-binom}

- $nobs_{t=t'}  \in \{0,1,2,3, ..., I\}$: number of observations of animals summed across all locations, at year $t'$
- $\psi_{t=t'}  \in [0,1]$: probability of presence at time $t = t'$
  - here assumed invariant across all locations
- $I$: the total number of locations

If we allow for **location-specific probabilities of presence**, the situation gets slightly more complicated.
In that case $nobs_{t=t'}$ could have been generated by the **Poisson Binomial distribution**, which captures a set of Bernoulli trials with unequal probabilities of success/presence.
There is no particular assumption made about how the different probabilities of presence arise (i.e., they are assumed to be fixed).
Importantly, the distribution can be approximated by a Binomial distribution, with a probability of presence corresponding to the mean probability of presence across locations (see @fig-plot-poisbinom).
This means that the following results, which hold exactly for the location-unspecific case, hold approximately for the location-specific case.
<!-- quantification of approximation error?-->
<!-- Beta Binomial distribution: https://stats.stackexchange.com/questions/390023/poisson-binomial-vs-beta-binomial]-->



```{r fig-plot-poisbinom}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Densities of the Poisson binomial (solid) vs. approximating Binomial distribution (dashed)"
#| out-width: 150%

# function to draws from the poisson binomial distribution
source("../../_functions/rpoisbinom.R")

# number of locations
I = 100

data_poisbin <- lapply(
  seq(1,5000),
  \(i){
    rpoisbinom(
      I = I,
      p_min = 0, p_max = 0.5,
      runif_seed = 123, rbinom_seed = i
      ) |>
      cbind(i = i)
  }
) |> dplyr::bind_rows()

# density poisson binomial
tmp <- data_poisbin$y_sum |> density()
dens_poisbin <- data.frame(
  x = tmp$x,
  y = tmp$y
)

# density binomial
dens_binom <- lapply(
  seq(0,I),
  \(i){
    data.frame(
      x = i,
      y = dbinom(x = i, size = I, prob = data_poisbin$psi_mean[1])
    )
  }
) |> dplyr::bind_rows()

ggplot2::ggplot(data = dens_binom, ggplot2::aes(x = x, y = y)) + 
  ggplot2::geom_line(linetype = "dashed") + 
  ggplot2::geom_line(data = dens_poisbin) + 
  ggplot2::theme_bw() + 
  ggplot2::labs(
    x = "Number of successes",
    y = "Probability"
  )

```

## Analysis setup

### Starting point rationale

Starting from a specific **target value for the probability of presence** (i.e., some sort of critical value from an ecological perspective; also the null hypothesis), we can investigate how the **minimal deviation detectable with a desired statistical power** changes with the sample size.
Moreover, we can specify a **critical minimal deviation**, we want to be able to detect and get the required sample size.

We investigate this in the context of the **Binomial test**.
The Binomial test can be used to determine whether a proportion of successes (sightings) in a number of Bernoulli trials (observed sampling locations) differs significantly from an expected proportion of successes (target value). 
The test statistic is the Binomial distribution (@eq-binom).

The following figure visualizes for 100 trials (sampling locations) how much statistical power we have to detect deviations of two different sizes from a certain target (h_a1 and h_a2, relative to h_0).
We can look at a whole range of sampling locations and each time determine the deviation that is detected with exactly the desired power.
This is the minimal detectable deviation given the desired power.


```{r fig-plot-binomial-power}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Power of the binomial test (I = 100)"
#| out-width: 150%

source("../../_functions/plot_binom_power.R")
plot_binom_power(
    n = 100, # number of trials
    psi_0 = 0.05, # probability of succes under h_0
    psi_a = c(0.1, 0.15), # probability of succes under h_a, or vector if multiple values
    alpha = 0.1 # type 1 error
)$plot_power
```


### Specific implementation

For the Chinese muntjac we have:

- target probability of presence: 0, as eradication is the goal
- critical minimal deviation: small, e.g., a probability of presence of 0.05 or 0.01; these mean that animals are expected to be present at 5 or 1 % of all possible sampling locations
- desired statistical power (i.e., the probability of detecting the critical minimal deviation): high, e.g., 0.90
- We rely on the one-sided Binomial test, as the probability of presence cannot be smaller than 0.

The following figure shows the minimal detectable deviation (mdd) as a function of the number of sampling locations (nl) (panel A), as well as the inverse, the number of sampling locations as a function of the minimal detectable deviation (panel B).


```{r fig-minimal-deviation}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Minimal detectable deviation and sample size (given power = 0.90)"
#| out-width: 150%
#| fig-width: 10
#| fig-height: 8

if (FALSE){
  source("../power_binom.R")
  plot_power + 
    ggplot2::labs(
      title = NULL
    )
}

# closed form expression one-sided CI
# can be used to find mdd / nl
get_mdd <- function(alpha, nl) {
  1 - ((alpha)^(1/nl))
}
get_nl <- function(mdd, alpha) {
  log(alpha)/log(1 - mdd)
}

# minimal detectable deviation
data_mdd <- data.frame(
  power = 0.9,
  nl = seq(1,600)
) |>
  dplyr::mutate(
    mdd = purrr::map2_dbl(nl, power, \(x, y){
      get_mdd(nl = x, alpha = 1 - y) 
    }),
    nl_hl = dplyr::case_when(
      nl %in% c(25, 50, 100, 250, 500) ~ nl,
      TRUE ~ NA_real_
    ),
    nl_lab = dplyr::case_when(
      !is.na(nl_hl) ~ sprintf("(nl = %.*f, mdd = %.*f)", 0, nl, 4, mdd),
      TRUE ~ NA_character_
    )
  )

# minimum number of locations
data_nl <- data.frame(
  power = 0.9,
  mdd = seq(from = 0.999, to = 0.0025, by = -0.001)
) |> 
  dplyr::mutate(
    nl = purrr::map2_dbl(mdd, power, \(x, y){
      get_nl(mdd = x, alpha = 1 - y) |> ceiling()
    }),
    mdd_hl = dplyr::case_when(
      round(mdd, digits = 4) %in% c(0.005, 0.01, 0.05, 0.1, 0.5) ~ mdd,
      TRUE ~ NA_real_
    ),
    mdd_lab = dplyr::case_when(
      !is.na(mdd_hl) ~ sprintf("(mdd = %.*f, nl = %.*f)", 4, mdd, 0, nl),
      TRUE ~ NA_character_
    )
  )
 
# visualize minimal detectable deviation as a function of sample size
plot_mdd <- ggplot2::ggplot(data = data_mdd, ggplot2::aes(x = nl, y = mdd)) + 
  ggplot2::geom_line() + 
  ggplot2::geom_point(ggplot2::aes(x = nl_hl)) + 
  ggplot2::geom_text(ggplot2::aes(label = nl_lab), 
                     hjust = 0.4, vjust = -0.6, size = 3) +
  ggplot2::labs(
    x = "Number of sampling locations",
    y = "Minimal detectable deviation"
    ) + 
  ggplot2::theme_bw()

# visualize sample size as a function of minimal detectable deviation
plot_nl <- ggplot2::ggplot(data = data_nl, ggplot2::aes(x = mdd, y = nl)) + 
  ggplot2::geom_line() + 
  ggplot2::geom_point(ggplot2::aes(x = mdd_hl)) + 
  ggplot2::geom_text(ggplot2::aes(label = mdd_lab), 
                     hjust = -0.1, vjust = -0.3, size = 3) +
  ggplot2::labs(
    x = "Minimal detectable deviation",
    y = "Number of sampling locations"
    ) + 
  ggplot2::theme_bw()


library(patchwork)
(plot_mdd + ggplot2::labs(title = "A: Minimal detectable deviation in function of sample size")) / 
  (plot_nl + ggplot2::labs(title = "B: Sample size in function of minimal detectable deviation")) + 
  patchwork::plot_layout(guides = "collect")


```

We can for instance see that 100 locations allow to detect a probability of presence of `r data_mdd |> dplyr::filter(nl == 100) |> dplyr::pull(mdd) |> round(x = , digits = 4)` and that a probability of presence of 0.01 requires `r data_nl |> dplyr::filter(round(mdd, digits = 4) == 0.01) |> dplyr::pull(nl)` sampling locations.


### Alternative rationale

Alternatively, we could argue that we simply consider the case in which we observe zero animals:
If at least one observation of an animal is made, action has to be taken anyway.
<!-- We now want to know how many zero-observations we need to being able rule out a critical probability of presence.-->
If zero observations are made across locations, however, we can guess that there are no animals
(i.e., the (maximum likelihood) point estimate of the probability of presence is zero).
The precision associated with this guess however depends on the number of sampling locations:
we will trust our guess more having 100 as compared to 10 locations with zero-observations.
The more locations without sightings, the less plausible probabilities of presence estimates larger than zero become.
We can formalize this by looking at the confidence interval (CI) estimate, which narrows for an increasing numbers of zero-sightings.
Values within the CI are alternative probability of presence values which are still plausible given the respective number of zero-sightings.^[Interpreting single CI estimates is tricky, and I need to double check whether this interpretation is admissible (enough); for now based on Lakens: https://lakens.github.io/statistical_inferences/07-CI.html]
We thus desire the sample size for which the interval does no longer include a critical minimal deviation from the value of zero probability of presence (see before).

It turns out that this way of approaching the question is not so different from the first way after all:
Relying on the inverse relationship between a CI and a statistical hypothesis test (ref: Casella & Berger),
we know that we would have rejected a zero probability of presence only if we had observed presences expected under probabilities outside the CI.^[This sounds a bit strange, but is due to the imperfect precision of our point estimate.] 
The upper bound of the CI thus marks *some* minimal detectable deviation from our best guess (zero probability of presence) corresponding to the earlier null hypothesis or target value.

Being able to reject a null hypothesis per se does not generally guarantee that we end up with the desired power.
In the case of zero probability of presence however as the null hypothesis, we come extremely close to the desired power once we reject the null hypothesis. @fig-minimal-deviation can thus also be interpreted as showing the upper CI bound under zero sightings instead of the minimal detectable deviation.^[I have explored this quasi-equality numerically for a range of values of $\alpha$ and $\beta$ (the type-II error). It is so evident, that I rely on the CI bound to produce @fig-minimal-deviation. The closed-form expression for the the one-sided CI under zero successes can be found in [@tony_cai_one-sided_2005]. To establish the equality more generally, one would need to show that $\alpha \approx \beta$, so that $(1 - \alpha) \approx (1- \beta)$, knowing that the CI always lies entirely to the right of zero.]


<!-- For > 58 locations, the CI estimate excludes a probability of presence of 5%, which might be a critical threshold from an ecological perspective. 
A probability of presence of 1% is however still included in the interval.
To exclude that value, we would need to measure at > 298 locations.-->


### Classification of observable effects

We can further complement the above approach(es) by taking estimation precision of all possible observable effects into account:
For each sample size, we can for all possible sample configurations (all possible numbers of sightings) obtain the observable effects (observable probability of presence) and their CI estimate.
Based on width and position of the CIs with respect to the target and the minimal detectable deviation, we can categorize each observable effect according to the [effectclass-logic](https://inbo.github.io/effectclass/articles/visualisation.html#format_ci-and-classification).
The following figure illustrates this and shows which interpretations we can afford given different sample sizes and observed probabilities of success.
Each point corresponds to an observable effect, its color codes its classification.
The minimal detectable deviations are marked via a black line.

<!--- fix threshold issue ---->
<!--- use points ---->


```{r fig-classification}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Classification of observable effects"
#| out-width: 150%

source("../plot_effectclass.R")

plot_ec

```




<!----------------------------------------------------------------------------------------------------------------------------------------------->
<!-- Scopes A & B ---------------------------------------------------------------------------------------------------------------------------->
<!----------------------------------------------------------------------------------------------------------------------------------------------->


# Scope B - data generation & analysis setup {.hidden}

::: {.callout-note icon=false}

**Summary**

[Summary here]

:::

## Data-generating model

### Data-generating distribution

[HERE adjust to scope A only]

Monitoring scopes A and B (abundance under management - status & trend) concern abundances of animals as outcomes (i.e., numbers of individuals per location), so we will rely on the **Poisson distribution** for these cases, with the parameter of focus being the expected number of observed animals:

$$
nobs_{i,d,t} \sim Pois(\mu_{i,t})
$$ {#eq-pois}
with

- $nobs_{i,d,t} \in \{0,1,2,3, ...\}$: number of observations of animals at location $i$, day $d$, year $t$
- $\mu_{i,t} \in (0, \infty)$: expected number of observations of animals at location $i$, year $t$ 
  - within a location and year, measurements from different days are treated as replicates, all independent realizations from the same Poisson distribution 
  - this reflects assumption of some stability over time (the alternative would be to assume variations across days and have day enter a regression model in some way (e.g., random effect, temporal trend of some shape))

When looking at status, we can assume we are looking at one specific year, so $t = t'$.

In the case of @eq-pois, we assume that $\mu_{i,t}$ is linked to the expected true population size / the true abundance at location $i$, year $t$, $\nu_{i,t}$.^[Following the logic of N-mixture modeling, or latent variable modeling, one would rather assume that $nobs_{i,d,t}$ is linked to the true population size at location $i$, year $t$, $n_{i,t}$; if the link holds at that level then it should also be implied at the level of expectations, but not the reverse.]
For simplicity, we assume a linear relationship that is invariant across all dimensions considered (i.e., space and time), but also across abundance itself:

$$
\mu_{i,t} = a * \nu_{i,t}
$$ {#eq-pois2}

In an ideal case $a = 1$, which means we can make statements about absolute abundances over space, time and abundance.
If $a \neq 1$, then we can still make statements about relative abundances over space, time and abundance *without having to know* $a$.
Reasons for $a \neq 1$ can be: an imperfect detection probability; animals being recorded more than once within a day. 

However, this assumption might be problematic if animals are very territorial, with no or little overlap between territories.
Then we are most likely only observing one to a few individuals and the number of observations is more related to their activity during that day.
In that case, we would need to find another approach to generate and analyse abundances.
One option could be the logistic regression model (see notes later), if abundances are basically one at this geographical scale, then we are dealing with data of presence and absence.

HERE: Here we have 2 problems: abundances rather 1; multiple observations more likely to pertain to the same animal(s) than to different ones.

If abundance at the given geographical scale is rather about the presence of animals, then we can alternatively rely on the **Bernoulli distribution** as the data-generating distribution. The parameter of focus then is the probability of presence:

$$
obs_{i,d,t} \sim Bern(\psi_{i,t})
$$ {#eq-bern}

- $obs_{i,d,t}  \in \{0,1\}$: observation of animals at location $i$, day $d$, year $t$
- $\psi_{i,t}  \in [0,1]$: probability of presence at location $i$, year $t$
  - within a location and year, measurements from different days are again treated as independent realizations from one Bernoulli distribution 
  - this again reflects assumption of some stability over time

In analogy to @eq-pois2, we can here think of the probability of presence featuring in @eq-bern as being an observed one, which, again, 
should be linked to a true probability of presence.
Ideally the link is a one-to-one link.
Again, an imperfect detection probability can weaken the link.
The same animals being present multiple times or multiple animals being present should not be an issue and can be accommodated by thinking of $\psi$ as the probability of at least one animal being present.

HERE
The first angle (@eq-bern1) makes sense if we want draw on the logistic regression model to include location-specific effects (and spatial dependencies) in our analysis.
The second angle (@eq-binom) is more of an aggregate approach.
We discuss the potential benefits of including location-specific effects later on.



### Regression model under monitoring scope B

#### Poisson regression

[HERE remove / integrate below]

In the case of both monitoring scopes A & B (abundance under management - status & trend), we can rely on the **Poisson regression model**. Starting from the distributional assumption in @eq-pois, we linearly decompose $\mu_{i,t}$, the parameter of the Poisson distribution, after applying the log-link function:

$$
log(\mu_{i,t}) = \eta_{i,t} \\
\Leftrightarrow \\
\mu_{i,t} = exp(\eta_{i,t})
$$ {#eq-poisreg1}
with:

- $log(\cdot)$: natural logarithm, here the log-link function
- $\eta_{i,t}$: linear predictor, which is some linear function of a set of predictor variables
- $exp(\cdot)$: exponential (i.e., inverse of log-link) function

The most simple model variant under monitoring scope B (abundance under management - trend) includes an intercept and a linear effect of year, both location unspecific, so index $i$ is dropped:

$$
\eta_{t} = \beta_{0} + \beta_{1} \, year_t^*
$$ {#eq-poisreg2}
with:

- $\beta_{0}$: intercept; linear predictor if $year_t^*= 0$
- $\beta_{1}$: slope associated with $year_t*$, which is some linear function of year

The **parameter of interest** - corresponding to our monitoring target 'trend' - is obviously the **slope**, $\beta_{1}$.
It's interpretation in terms of the linear predictor, $\eta$, is the one we are familiar with from simple linear regression:
$\beta_1$ reflects the absolute change in linear predictor values given a one-unit change in year*.

The interpretation in terms of the expected number of observed animals, $\mu$, needs to incorporate the link-function used:
Here, $exp(\beta_1)$ reflects the relative difference (ratio) in expected numbers of observations given a 1-unit difference in year*.
We can derive this as (leaving out subscript $t$):

\begin{align}
log(\mu)|(year^*_1 + 1) - log(\mu)|year^*_1 &= \beta_1 \quad\quad|\;exp(\cdot) \\
exp(log(\mu)|(year^*_1 + 1) - log(\mu)|year^*_1) &= exp(\beta_1) \\
\frac{exp(log(\mu)|(year^*_1 + 1))}{exp(log(\mu)|year^*_1)} &= exp(\beta_1) \\
\frac{\mu|(year^*_1 + 1))}{\mu|year^*_1} &= exp(\beta_1)
\end{align}

So, $exp(\beta_1)$ is the factor by which $\mu$ changes given a one-unit change in year
($\mu|(year^*_1 + 1) = exp(\beta_1)*\mu|year^*_1$),
and $(exp(\beta_1)-1)*100$ is the percentage change in $\mu$ associated with a one-unit change in year.

The following figure shows a number of different trends implied by different sizes of $\beta_1$ for both $\eta$ and $\mu$.

```{r fig-size-trend}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Visualization of different effect sizes of beta_1 in the Poisson regression model"
#| out-width: 150%

data_trend <- data.frame(
  # size of linear effect of year
  effect_size = -seq(0, 2, 0.2)
) |>
  tidyr::crossing(
   year_star = seq(0,15)
) |>
  dplyr::mutate(
    eta =  effect_size*year_star + 2,
    mu = exp(eta),
    effect_size = factor(effect_size) |> forcats::fct_rev()
    ) |>
  tidyr::pivot_longer(
    cols = tidyselect::all_of(c("eta", "mu")),
    names_to = "out_name",
    values_to = "out_value"
  )
ggplot2::ggplot(
  data = data_trend, 
  mapping = ggplot2::aes(x = year_star, y = out_value, group = effect_size)
  ) + 
  ggplot2::geom_line(ggplot2::aes(color = effect_size), linewidth = 1.2) +
  ggplot2::facet_wrap(~ out_name, scales = "free_y", ncol = 2) + 
  ggplot2::labs(x = "year*", y = "outcome", color = "size beta1") +
  ggplot2::scale_x_continuous(breaks = data_trend$year_star |> unique()) + 
  ggplot2::scale_colour_viridis_d(direction = -1) + 
  ggplot2::theme_bw()

```

In case $\beta_1 = 0$, we have `r round((exp(0)-1)*100, digits = 0)` %,
if $\beta_1 = -0.2$, we have around `r round((exp(-0.2)-1)*100, digits = 0)` %, 
if $\beta_1 = -0.4$, we have around `r round((exp(-0.4)-1)*100, digits = 0)` %, 
if $\beta_1 = -1$, we have around `r round((exp(-1)-1)*100, digits = 0)` % decrease per year. 

What could be an interesting effect size from a biological perspective?

#### Logistic regression

Alternatively, we can start from the distributional assumption made in @eq-bern and rely on the **Logistic regression model** 
and we linearly decompose $\psi_{i,t}$, the parameter of the Bernoulli distribution, after applying the logit-link function:


$$
logit(\psi_{i,t}) = log(\frac{\psi_{i,t}}{1-\psi_{i,t}}) = \eta_{i,t} \\
$$ {#eq-logreg1}
with:

- $logit(\cdot)$: logit-link function (i.e., the log-odds/ logarithm of the odds)
- $log(\cdot)$: natural logarithm
- $\frac{\psi_{i,t}}{1-\psi_{i,t}}$: odds of presence
- $\eta_{i,t}$: linear predictor

The most simple model variant is the one from above (@eq-poisreg2), with the parameter of interest again being the **slope**, $\beta_{1}$.
This parameter can again be interpreted in terms of the linear predictor,
in terms of the odds of presence ($odds_{i,t}: exp(\eta_{i,t}) = \frac{\psi_{i,t}}{1-\psi_{i,t}}$),
or in terms of the probability of presence ($\psi_{i,t} = \frac{odds_{i,t}}{1+odds_{i,t}}$)

The interpretation in terms of the odds of presence is analoguous to the interpretation above:
So, $exp(\beta_1)$ is the factor by which the odds of presence change given a one-unit change in year
The following figure shows a number of different trends implied by different sizes of $\beta_1$ for $\nu$ (the log-odds), the odds and $\psi$.

```{r fig-size-trend-2}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Visualization of different effect sizes of beta_1 in the logistic regression model"
#| out-width: 150%

data_trend <- data.frame(
  # size of linear effect of year
  effect_size = -seq(0, 2, 0.2)
) |>
  tidyr::crossing(
   year_star = seq(0,15)
) |>
  dplyr::mutate(
    eta =  effect_size*year_star + 2,
    odds = exp(eta),
    psi = odds / (1 + odds),
    effect_size = factor(effect_size) |> forcats::fct_rev()
    ) |>
  tidyr::pivot_longer(
    cols = tidyselect::all_of(c("eta", "odds", "psi")),
    names_to = "out_name",
    values_to = "out_value"
  )
ggplot2::ggplot(
  data = data_trend, 
  mapping = ggplot2::aes(x = year_star, y = out_value, group = effect_size)
  ) + 
  ggplot2::geom_line(ggplot2::aes(color = effect_size), linewidth = 1.2) +
  ggplot2::facet_wrap(~ out_name, scales = "free_y", ncol = 3) + 
  ggplot2::labs(x = "year*", y = "outcome", color = "size beta1") +
  ggplot2::scale_x_continuous(breaks = data_trend$year_star |> unique()) + 
  ggplot2::scale_colour_viridis_d(direction = -1) + 
  ggplot2::theme_bw()

```


If $\beta_1 = -0.2$, then the odds change by a factor of `r round(exp(-0.2), digits = 2)` per year, 
and if $\beta_1 = -1$, that factor is `r round(exp(-1), digits = 2)`. 

The question is again, what could be an interesting effect size from a biological perspective?


## Analysis setup


# Scope A - data generation & analysis setup {.hidden}

::: {.callout-note icon=false}

**Summary**

[Summary here]

:::

## Data-generating model

### Data-generating distribution

[HERE: reduce to data-generating distribution same as detection]

### Regression model 

For monitoring scope A (abundance under management - status), we could consider a special case of the model specified for monitoring scope B. 
Specifically, we could further simplify the above regression model(s) (@eq-poisreg2) to an intercept-only model to estimate expected local population size / probability of presence at year $t=t'$.
The following thoughts on analysis setup for monitoring scope B then apply analogously to scope A.

Alternatively, we might instead of a somewhat artificial regression model just use a Poisson or Binomial test (see later elaborations).


## Analysis setup

Based on the simple model specified above (@eq-poisreg2; analogue considerations hold for the logistic regression variant), we can now either

- determine a few minimal trend sizes of interest, and analyse how statistical power to detect those effects varies with the number of locations 
- determine a few interesting numbers of locations, and analyse how the size of the minimal effect that can be detected with a certain statistical power varies

In addition, we can also extend the above data-generation model in a number of ways, the most obvious ones being

- the inclusion of location-specific effects (e.g., a random intercept and/or a random slope) 
- the accommodation of a non-perfect detection probability

Such additional complexities could be included in both the data-generating and analysis model (case 1),
only the data-generating model (case 2),
only the analysis model (case 3, which we discard as not realistic),
or in none of the models (case 4, which we are in now).
In case 1, we expect that simulating and modeling additional complexities will lead to additional estimation variance,
which will lead to larger sample sizes being required to keep statistical power at the desired level.
In case 2, we expect that simulating but not modeling additional complexities will most often lead to estimation bias, which will also affect power, albeit in less predictable ways. 
Investigating the effects of missspecification is also a bit of a different question and should maybe not be of central interest for the present power analysis (or maybe it should).
However, for a relevant special case, neglected imperfect detection probability, we can form an expectation:
Here a trend will like be underestimated lowering statistical power to detect it.
In general, analyses under cases 1 and 2 will reflect more realistic scenarios, but the simple case 4 can function as a lower bound estimate (i.e., it probably won't get better than this). 

Another direction for extensions concerns sample size as multi-dimensional:
Until now it was implicit that we assumed the number of days to be fixed (e.g., at 30; cf. @bollen_gebruik_nodate), while they could also be varied.
In the simple model above, we should see a one-to-one trade-off effect of days and locations (adding one day at one location should have the same effect as adding one location for one day), which is why we can keep days as well fixed.
This situation changes when costs are considered, where different dimensions can have very different contributions to the overall costs (e.g., leaving an existing camera trap in the field for some more time is cheaper than placing an additional new one).
So, if costs become an additional optimization criterion next to statistical power, then it would make sense to vary sample size along multiple dimensions to investigate trade-offs.
(Also when location-specific effects and/or detection probabilities are estimated, sample size dimension trade-off effects will become more interesting and should probably be added to the picture.) 





<!----------------------------------------------------------------------------------------------------------------------------------------------->
<!-- Sampling ----------------------------------------------------------------------------------------------------------------------------------->
<!----------------------------------------------------------------------------------------------------------------------------------------------->


# Sampling

::: {.callout-note icon=false}

**Summary**

[Summary here]

:::


Thinking about sampling answers the question 
"How do (simulated) data map onto geographical locations?".

This is needed to visualize data collection under different scenarios,
helps judging whether assumptions during data generation are more or less realistic, and 
helps determining the range of realistic sample sizes.

The following sections in this chapter deal with

- the definition of a sample frame
- the drawing of a random sample of locations
- the mapping of future simulated data on the random sample of locations


## Defining the sample frame

### Defining the distribution area

All above monitoring scopes concern the currently known **distribution area** in the province of Antwerp, 
including the municipalities of Schoten, Brasschaat, and Wuustwezel. 
To define the sample frame, we thus rely on the [Provisional Reference File for Municipal Boundaries](https://www.vlaanderen.be/datavindplaats/catalogus/voorlopig-referentiebestand-gemeentegrenzen-toestand-01-01-2025-correctie-11-02-2025) published by the Flemish government agency "Digital Flanders" (ref). 
This file defines the boundaries of the administrative units in the Flemish Region, 
and provides us with the geometry of the here-relevant Flemish municipalities. 

Alternatively, recent reliable observations could be used to derive the distribution area.


### Defining a buffer around the distribution area

While some monitoring scopes focus on the distribution area, others concern the detection of animals around the distribution area.
Therefore, a buffer area around the distribution area must be defined.
Here, we rely on estimates of the species' mobility / action radius and thus take a buffer of [2000 m]{style="background-color: #E6E6FA;"} width around the distribution area.

[(Recap information on the species' mobility here and adapt width.)]{style="background-color: #E6E6FA;"}


```{r fig-plot-dist-buff-area}
#| include: true
#| message: false
#| fig-cap: "Map of distribution (inner) and buffer (outer) area (both in yellow)." 
#| out-width: 150%



mapview::mapview(
    list(
      maplist_3857$map_antw_merged,
      maplist_3857$map_buff
    ),
    col.regions = list(INBOtheme::vl_darkyellow, INBOtheme::vl_darkyellow),
    alpha.regsions = 0.3,
    color = INBOtheme::vl_darkred
  )

 
```



### Selecting habitat areas in the distribution and buffer area

Within both the distribution and buffer area, we require geometrical information on **(potential) habitat areas** of the Chinese muntjac to define the sample frame.
We know that the species can generally be found in terrestrial habitats with dense vegetation, mainly forests/woodlands, forest edges, hedgerows, abandoned plantations, peri-urban overgrown parks and gardens, and patches of natural vegetation in agricultural areas.
[Based on habitat maps, which habitat type most frequent in Flanders?]{style="background-color: #E6E6FA;"}
We thus use OpenStreetMap data (ref) extracting relevant features via suitable [vegetation](https://wiki.openstreetmap.org/wiki/Vegetation) keys and tags. 

OpenStreetMap key-tag combinations currently used are:

- 'natural = wood' & 'landuse = forest' to obtain (managed and unmanaged) forest ([ref](https://wiki.openstreetmap.org/wiki/Forest))
- 'landuse = orchard' to obtain tree and scrub plantings ([ref](https://wiki.openstreetmap.org/wiki/Tag:landuse%3Dorchard))

[This list can be extended (e.g., by adding key-tag combinations such as 'natural=scrub') or refined (e.g., by adding keys specifying forest type; by considering distance to other features such as cities or open landscape).]{style="background-color: #E6E6FA;"}

To select those (clusters of) habitat areas with a certain minimal size, the information obtained via OpenStreetMap is then processed in the following way:
[(Describe processing here.)]{style="background-color: #E6E6FA;"}
Within the buffer area, habitat areas are additionally required to have a maximal distance of [2000 m]{style="background-color: #E6E6FA;"} (see reasoning above) to the closest habitat area within the distribution area to be selected.

The habitat areas remaining within the distribution area form the **sample frame** for the distribution area and hence monitoring scopes A, B, and D.
[(Add that criterion is minimal overlap; sampling frame can thus extend beyond municipaltity borders.)]{style="background-color: #E6E6FA;"}

For monitoring scope C targeting a buffer around the distribution area, the sample frame is formed by the habitat regions remaining within the buffer area and not selected for the sampling frame of the habitat area. 


```{r fig-plot-sample-frame}
#| include: true
#| message: false
#| fig-cap: "Map of sample frame for distribution (light green) and buffer area (dark green)." 
#| out-width: 150%

mapview::mapview(
    list(
      maplist_3857$map_antw_hab,
      maplist_3857$map_buff_hab
    ),
    col.regions = list(INBOtheme::vl_lightgreen, INBOtheme::vl_darkgreen)
  )

 
```


## Drawing a random sample {.hidden}

HERE. 
Draw a spatially-balanced random sample using a variant of the GRTS-algorithm as implemented in the grtsdb R package.

Steps: 

Tessellation:
With respect to a bounding box overlaying both distribution and buffer area.
Define cell size.
Overlay distribution / buffer area with spatial grid consisting of quadratic cells of dimension $p \times p$ m [here $100 \times 100$ m for illustrational purposes]

Draw balanced sample

Discard grid cells outside sample frame
(minimum overlap)

Keep first N cells.



```{r fig-plot-sample}
#| include: true
#| message: false
#| fig-cap: "Map of sample for distribution area combined (red)." 


source("../grts_sample.R")

mapview::mapview(
    list(
      maplist_3857$map_antw_hab,
      sample_hab_upd
    ),
    col.regions = list(INBOtheme::vl_lightgreen, INBOtheme::inbo_oranjerood),
    alpha.regions = list(0.2, 0.9)
  )

 
```


HERE
For this illustration, we draw a naive random sample of locations.
To this end, we take the following steps:



```{r sampling-create-grid}
#| include: false
#| cache: false
#| results: 'hide'

# create grid over bounding box around distribution area
p = 100
grid_box <- sf::st_make_grid(
  maplist_3857$map_antw_merged,
  cellsize = p # meters, since Belgian Lambert
) |>
  sf::st_sf()
if (FALSE) mapview::mapview(list(maplist_3857$map_antw_merged, grid_box))

# extract grid cells that overlap with habitat polygons
grid_hab <- sf::st_filter(
  grid_box,
  maplist_3857$map_antw_hab
  #,.predicate = sf::st_within
)
if (FALSE) mapview::mapview(list(map_mun_antw_hab, grid_hab))

# extract grid cells that overlap with habitat polygons
# https://github.com/r-spatial/sf/issues/1125


```

2. Draw random sample of $C$ grid cells extracted in the previous step (here $100$ grid cells)

```{r sampling-draw-random-sample}
#| include: false
#| eval: false
#| cache: false
#| results: 'hide'

# --- draw naive random sample ---------------------------------------------

# sample size
ncell <- 500
assertthat::assert_that(ncell < nrow(grid_hab))

# sample
grid_sample <- grid_hab |>
  dplyr::sample_n(tbl = _, size = ncell)

# GRTS
# https://inbo.github.io/grtsdb/articles/basic.html
# implementation?

```



Within each sampled grid cell, a camera trap is assumed to be placed following a certain strategy (e.g., placement in center or corner of grid cell). 

The total number of grid cells associated with the sample frame forms the upper bound of sample sizes to consider in a power / precision analysis.


##### Specifications: Sampling

- Determine grid
  - cell size?
    - [ ] value from literature (250 m; cf. @bollen_gebruik_nodate)
    - [ ] value based on camera angle/radius (usually $< 10$ m?)
    - [x] smallish value to keep things visible
    - [ ] value based on habitat parameters (e.g., forest density; to make sure a grid cell contains suitable areas for camera placement)
  - how to define intersection with habitat areas?
    - [x] grid cell overlaps with habitat areas 
    - [ ] overlap at least $q \gg 0$ %
- Draw a random sample
  - strategy?
    - [x] naive random sample
    - [ ] spatially balanced random sample via the GRTS method (INBO R package *grtsdb*)







# References

::: {#refs}
:::
