---
title: "Illustration power simulation for chinese muntjac"
format: 
  html:
    toc: true
    toc-location: right-body
    toc-depth: 3
    toc-expand: true
    theme: yeti
    number-sections: true
    number-depth: 3
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
lightbox: true
bibliography: ../../../media/MIAS.bib
execute: 
  eval: true
  echo: true
  cache: false
---

The following documents deals with 2 monitoring scopes:

- Abundance under management: Track (changes in) population size in relation to sustained eradication measures in distribution area
- Detection under management: Detect potential spread into buffer area around distribution area in relation to sustained eradication measures in distribution area

For now we mainly focus on the first scope.

The document summarizes some thoughts on the following components of a simulation-based power analysis:

- Sampling: How could data look like?
- Data-generation: What is an as-complex-as-necessary / as-simple-as-possible data-generating model?
- Modeling: What is an as-complex-as-necessary / as-simple-as-possible data-analysis model?
- Simulation: Initial thoughts on design and implementation


## Sampling


### Define the sample frame

To define the **sample frame** per monitoring scope, we take the following 2 preparatory steps:

1. Get map layers: 
  - Flanders' municipalities (as distribution is currently tied to specific municipalities)
  - Flanders' habitat area (e.g., forest)


```{r sampling-get-map-layers}
#| include: true
#| results: 'hide'

# --- get map layers---------------------------------------------

# municipalities
map_mun <- inbospatial::get_feature_wfs(
  wfs = "https://geo.api.vlaanderen.be/VRBG/wfs",
  layer = "VRBG:Refgem"
)
# https://www.vlaanderen.be/datavindplaats/catalogus/
# voorlopig-referentiebestand-gemeentegrenzen-
# toestand-01-01-2025-correctie-11-02-2025
#
# check CRS
if (FALSE) sf::st_crs(map_mun)$Name # EPSG:31370

# habitat
map_hab <- sf::st_read("../../../data/gis/inbo_boskartering/bosvl2001.shp")
# get forest map (vector?)
# raster: https://www.vlaanderen.be/datavindplaats/catalogus/
# digitale-boswijzer-vlaanderen-2021
# raster: https://www.vlaanderen.be/datavindplaats/catalogus/
# groenkaart-vlaanderen-2021
#
# check CRS
if (FALSE) sf::st_crs(map_hab)$Name # EPSG:31370


```

2. Process map layers:
  - Select relevant municipalities (i.e., distribution area): Schoten, Brasschaat, Wuustwezel
  - Get buffer area of width $u$ meters around distribution area (here $u = 2000$)
    - Alternative: Get buffer area around habitat regions within distribution area?
  - Get habitat regions within distribution area (e.g., Forests)
  - Get habitat regions within buffer area

```{r sampling-process-map-layers}
#| include: true
#| cache: true
#| results: 'hide'
#| warning: false

# TO DO: check warnings

# --- process map layers ---------------------------------------------

# select relevant municipalities
map_mun_antw <- map_mun |>
  dplyr::filter(
    grepl("Schoten|Brasschaat|Wuustwezel", NAAM)
  )

# extract polygons from multisurface to enable further processing
# (mapview and intersection)
lapply(1:nrow(map_mun_antw),
       function (i) {
         assertthat::are_equal(
           sf::st_as_text(map_mun_antw$SHAPE[i]) |> grep("POLYGON", x = _),
           1
         )
         })
map_mun_antw <- sf::st_cast(map_mun_antw, "GEOMETRYCOLLECTION") |>
  sf::st_collection_extract("POLYGON")
if (FALSE) mapview::mapview(map_mun_antw)

# create buffer around municipalities
u = 2000
map_mun_antw_merged <- map_mun_antw |>
  dplyr::summarise()
tmp_buff <- map_mun_antw_merged |>
  sf::st_buffer(
    dist = u # meters, since Belgian Lambert
    )
map_mun_antw_buff <- sf::st_difference(
  tmp_buff,
  map_mun_antw_merged
)
if (FALSE) mapview::mapview(map_mun_antw_buff)

# quickly check municipality maps
if (FALSE) {
  mapview::mapview(
    list(map_mun_antw, map_mun_antw_buff),
    col.regions = list("blue", "red"),
    alpha.regions = .3
  )
}

# intersect habitat with municipality maps
map_hab <- sf::st_transform(x = map_hab, crs = "EPSG:31370")
map_mun_antw_hab <- sf::st_intersection(
  map_mun_antw,
  map_hab
  )
mapview::mapview(map_mun_antw_hab)
map_mun_antw_buff_hab <- sf::st_intersection(
  map_mun_antw_buff,
  map_hab
)
if (FALSE) mapview::mapview(map_mun_antw_buff_hab)

```

After these 2 steps we get the following maps.

```{r fig-plot-dist-buff-area}
#| include: true
#| cache: true
#| results: 'hide'
#| message: false
#| fig-cap: "Map of distribution and surrounding buffer area (both in yellow) and enclosed habitat regions (both in green). The distribution area is highlighted in lighter colors." 
#| out-width: 150%
#| layout-ncol: 1 

# transform: pseudo-mercator, which is CRS of map tile 
# (unless map tile - in pseudo-mercator - is first layer)
mapnames <-   c(
    "map_mun_antw",
    "map_mun_antw_hab",
    "map_mun_antw_buff",
    "map_mun_antw_buff_hab",
    "map_mun_antw_merged"
  )
maplist <- lapply(
  mapnames,
  \(i){
    tmpmap <- sf::st_transform(x = get(i), crs = "EPSG:3857")
  }
) 
names(maplist) <- paste0(mapnames, "_3857")
# check transformation
lapply(maplist, \(x) sf::st_crs(x)$Name)

# base plot
plot_base <- ggplot2::ggplot(
  data = maplist$map_mun_antw_3857 # warning if wgs84 is used
) +
  ggspatial::annotation_map_tile(
    type = "cartolight", zoom = 12, cachedir = tempdir(), alpha = .5 
  ) 
plot_names <- ggplot2::geom_sf_text(
    ggplot2::aes(
      label = NAAM
    ),
    size = 5,
    color = INBOtheme::vl_black,
    check_overlap = TRUE
  )

# plot distribution area
geoms_dist <- list(
  ggplot2::geom_sf(fill = INBOtheme::vl_yellow, 
                   alpha = .6),
  ggplot2::geom_sf(data = maplist$map_mun_antw_hab_3857, 
                   fill = INBOtheme::vl_lightgreen, 
                   color = INBOtheme::vl_lightgreen,
                   alpha = .6) 
)
plot_dist <- plot_base + geoms_dist + plot_names

# plot buffer area
geoms_buffer <- list(
    ggplot2::geom_sf(data = maplist$map_mun_antw_buff_3857, 
                     fill = INBOtheme::vl_darkyellow, 
                     alpha = .6),
  ggplot2::geom_sf(data = maplist$map_mun_antw_buff_hab_3857, 
                   fill = INBOtheme::vl_darkgreen, 
                   color = INBOtheme::vl_darkgreen,
                   alpha = .6)
)
plot_buff <- plot_base + geoms_buffer + plot_names

# combined plot
plot_dist_buff <- plot_base + geoms_dist + geoms_buffer +   plot_names +
  # emphasize dist
  ggplot2::geom_sf(data = maplist$map_mun_antw_merged_3857, 
                   fill = NA, 
                   color = INBOtheme::vl_black, 
                   linewidth = 1) 
plot_dist_buff

```


#### Monitoring scope 'abundance under management'

On the basis of these, we define the **sample frame for the monitoring scope 'abundance under management'** as the habitat regions within the distribution area.

We focus on the first monitoring scope in the remainder of this document.

#### Monitoring scope 'detection under management'

The **sample frame for the monitoring scope 'detection under management'** is defined as the habitat regions within the buffer area.  



### Draw a sample

For this illustration, we draw a **naive random sample**.
To this end, we take the following steps:

1. Tessellation of the sample frame:
  - Overlay distribution area with spatial grid consisting of quadratic cells of dimension $p \times p$ meters [here $p = 250$; cf. @bollen_gebruik_nodate]
  - Extract grid cells that show $q%$ overlap with habitat regions within distribution area (here $q > 0$) 

```{r sampling-create-grid}
#| include: true
#| cache: true
#| results: 'hide'

# create grid over bounding box around distribution area
p = 250
grid_box <- sf::st_make_grid(
  map_mun_antw_merged,
  cellsize = p # meters, since Belgian Lambert
) |>
  sf::st_sf()
if (FALSE) mapview::mapview(list(map_mun_antw_merged, grid_box))

# extract grid cells that overlap with habitat polygons
grid_hab <- sf::st_filter(
  grid_box,
  map_mun_antw_hab
  #,.predicate = sf::st_within
)
if (FALSE) mapview::mapview(list(map_mun_antw_hab, grid_hab))

```

2. Draw random sample of $C$ grid cells extracted in the previous step (here $C = 100$)

```{r sampling-draw-random-sample}
#| include: true
#| cache: true
#| results: 'hide'

# --- draw naive random sample ---------------------------------------------

# sample size
ncell <- 100
assertthat::assert_that(ncell < nrow(grid_hab))

# sample
grid_sample <- grid_hab |>
  dplyr::sample_n(tbl = _, size = ncell)

# GRTS
# https://inbo.github.io/grtsdb/articles/basic.html
# implementation?

```

The following plot visualizes the grid and the sample drawn.

```{r fig-plot-sample}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Visualization of the sampled (filled) vs. not sampled (empty) grid cells"
#| out-width: 150%

# transform: pseudo-mercator
grid_hab_3857 <- sf::st_transform(x = grid_hab, crs = "EPSG:3857")
grid_sample_3857 <- sf::st_transform(x = grid_sample, crs = "EPSG:3857")



# sample from sampling frame distribution area
plot_dist_sample1 <- plot_dist +
  ggplot2::geom_sf(data = grid_hab_3857, color = INBOtheme::vl_lightred, alpha = .3) +
  ggplot2::geom_sf(data = grid_sample_3857, color = NA, fill = INBOtheme::vl_lightred) +
  plot_names

plot_dist_sample1

```


3. Select location in sampled grid cells (not clear how this was done in e.g. @bollen_gebruik_nodate;
"Vervolgens plaatsten we in elk van de geselecteerde cellen twee keer per jaar gedurende telkens één maand een camera: één keer in de lente-zomer en één keer in de herfst-winter. Op die manier verzamelen we er maandelijks 32 deployments.")

A better alternative to drawing a naive random sample would be to e.g. use the GRTS-method to draw a **spatially balanced random sample**.

Stratification is an additional aspect that is however not considered in this illustration.


::: {.callout-note icon=false}
### ?
Applications that use R package *grtsdb* in a similar setting to draw a sample?
:::


## Data generation

### Data-generating model

We thinking about a possible data-generating model in the context of the generalized linear mixed modeling framework.
Also, whether the model is feasible to implement as a data-analysis model matters, to not complicate things.

For the variables occuring in the model we have the following dimensions:

- $c = 1, ..., C$ grid cells 
- $i = 1, ..., I$ sampling locations (here $i = c$ for all $i$ and $c$, so dimensions not separable; index $c$ left out for readability)
- $d = 1, ..., D$ days per measurement season (more precisely: $d_{s,t}$; here the same for all $t$)
- $s = 1, ..., S$ measurement seasons per year (more precisely: $s_t$; here the same for all $t$)
- $t = 1, ..., T$ years


#### Local population size - true vs. observed

For a start we have:

- $\nu_{i,d,s,t}$: *true* population size at location $i$, day $d$, season $s$, year $t$ (in the following also 'local population size' in short)
- $n_{i,d,s,t}$: observed population size at location $i$, day $d$, season $s$, year $t$ (in the following also 'observed local population size' i short)

In the simplest case, we can directly observe this variable, so that

$$
n_{i,d,s,t} = \nu_{i,d,s,t} 
$$

In a more realistic scenario, we cannot directly observe this variable, so, it can be that $n_{i,d,s,t} \neq \nu_{i,d,s,t}$.
Attempting to link the observed to the true local population size, we can for instance assume $n_{i,d,s,t}$ to be the number of successes in a sequence of $\nu_{.}$ Bernoulli trials with success (here: detection) probability $p$ (cf. "N-mixture" models; @madsen_review_2023):

$$
n_{i,d,s,t} \sim Binom(\nu_{.}, p)
$$

It is usually assumed that the true local population size shows some short-term stability, e.g., is invariant across days (i.e., $\nu_{.} = \nu_{i,s,t}$, so index $d$ is dropped; 'population closure' in time).
Other forms of stability should be possible (e.g., $\nu_{.}$ changes over days, but with strong carry-over effects), but also complicate things.

Implied seems also the assumption that $n_{i,d,s,t}$ does not contain repeated observations of the same animals (FIX: for a given day).
So, within a day, every observation is assumed to come from a distinct individual.
(Or: $n_{i,d,s,t} < \nu_{i,s,t}$ only because of $p < 1$, but not, because individuals are observed multiple times.)

::: {.callout-note icon=false}
### Assumptions

- time-invariance within seasons: plausible enough if measured before/after management action?
- unique observations within days: plausible? 
  - a-priori correction possible in theory, but difficult in practice, as muntjac are not 'naturally marked'
:::

In addition, a stable (i.e., time- and location-invariant) detection probability $p$ is assumed.
Seems not too implausible to begin with, although these stability restrictions can probably again be relaxed to some extent.

#### Linking local population size to its expectation

Following the GLM-logic, we assume that the local population size represents a draw from the Poisson distribution, i.e.:

$$
\nu_{i,s,t} \sim Pois(\mu_{i,s,t})
$$ {#eq-2}

with 

- $\nu_{i,s,t}$: population size at location $i$, season $s$, year $t$ (invariant across days)
  - the total population size across all locations would be: $N_{s,t} = \sum_{i = 1}^I\nu_{i,s,t}$ (FIX: move up together with assumption)
- $\mu_{i,s,t}$: expected population size at location $i$, season $s$, year $t$ (invariant across days)

Given that the total population size is calculated as the sum of the local population sizes, these need to be independent.
In other words, there is no overlap between locations, observations are assumed to come from distinct individuals between locations ('population closure' in space).

::: {.callout-note icon=false}
### Assumption
- no population-overlap between locations: plausible enough?
  - a-priori correction based on expected size of range of individuals?
:::


Possible extensions / alternatives for @eq-2 might account for:

- over-dispersion if, e.g., the negative binomial distribution is used instead of the Poisson distribution
- zero-inflation if we assume there is an additional process (rather than just small expected population sizes) producing zeros 


#### Decomposing / predicting the expected local population size

In the GLM-context, we assume:

$$
log(\mu_{i,s,t}) = \eta_{i,s,t}
$$
with:

- $log(\cdot)$: log-link function
- $\eta_{i,s,t}$: linear predictor

The linear predictor is assumed to be linearly decomposable into a number of effects.
In our case the following equations form a starting point.

At level 1 (within locations) we assume:

$$
\eta_{i,s,t} = \beta_{0,i} + \beta_{1,i} \, \text{season*} + \beta_{2, i} \, \text{year*}
$$

At level 2 (between locations) we assume:

\begin{align}
\beta_{0,i} &= \gamma_{00} + u_{0i}  \\
\beta_{1,i} &= \gamma_{10} + u_{1i}  \\ 
\beta_{2,i} &= \gamma_{20} + u_{2i}  \\
\end{align}


with:

- $\gamma_{00}$: fixed part of intercept; mean of linear predictor across locations if $season^*$ and $year^*$ are zero
- $ u_{0i}$: random part of intercept capturing locations-specific deviations from the fixed intercept
- $\gamma_{10}$: fixed part of effect of $season^*$ (mean effect across locations), which is some deterministic function of season (i.e., $season^* = g(season)$)
- $ u_{1i}$: random part of effect of $season^*$ capturing locations-specific deviations from the above fixed slope
- $\gamma_{20}$: fixed part of effect of $year^*$, which is some deterministic function of year (i.e., $year^* = h(year)$)
- $ u_{2i}$: random part of effect of $year^*$ capturing locations-specific deviations from the above fixed slope

Random effects/components are usually assumed to be multivariate normally distributed with zero mean vector and valid covariance matrix. 

::: {.callout-note icon=false}
### Terminology...

...can be confusing between different literatures, it seems.
For instance, year is not included as a random effect, rather as having a random effect, varying across locations.
The random effect of year random not in the sense of varying across years, but in the sense of varying across locations.
Year is a fixed predictor.
...

:::

Including an effect of season allows to have seasonal effects that deviate from the long-term trend (e.g., more individuals in autumn than winter/effect of winter).
It should be distinguishable from the intercept and effect of year once there are more than 2 years of data (?).

For the long-term trend, different shapes are possible in principle:
- linear? 
- exponential decay? 
- other non-linear trends? 
- mixture of trends (population growth and decline due to control; but control adapted to momentary population size)

The following figure illustrates a few shapes for the linear predictor as well as for the expected local population size (i.e., for $log(\mu_{i,s,t})$ and $\mu_{i,s,t}$).

```{r fig-shape-trend}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Visualization of different shapes of long-term trends"
#| out-width: 150%


data_trend <- data.frame(
  shape = c(
    "linear", 
    "exponential decay", 
    "doubly exponential decay"
    )
) |>
  tidyr::crossing(
   t = seq(1,10)
) |>
  dplyr::mutate(
    eta = 
      dplyr::case_when(
        shape == "linear" ~ -.9 * t + 10,
        shape == "exponential decay" ~ exp(-.4 * t),
        shape == "doubly exponential decay" ~ exp(exp(-.4 * t)),
          ),
    mu = exp(eta)
    ) |>
  tidyr::pivot_longer(
    cols = tidyselect::all_of(c("eta", "mu")),
    names_to = "out_name",
    values_to = "out_value"
  )

ggplot2::ggplot(
  data = data_trend, 
  mapping = ggplot2::aes(x = t, y = out_value, group = out_name)
  ) + 
  ggplot2::facet_wrap(shape ~ out_name, scales = "free_y", ncol = 2) + 
  ggplot2::geom_line() +
  ggplot2::theme_bw()

  

```


::: {.callout-note icon=false}
### Non-linearity
To stay within the GLM framework (holding model-estimation in mind), the model for the linear predictor should be linear in parameters. 
:::



Possible extensions:

- temporal dependencies taking temporal distance into account; carry-over effects in time
- spatial dependencies taking spatial distance into account; carry-over effects in space


::: {.callout-note icon=false}
### ?
How are such carry-over effects in space / time usually implemented, e.g. in *brms*?
:::


### Mapping data on sampled locations

- ...

## Modeling

### Analysis model

Some thoughts:

- start from data-generating model, but possibly simplify (e.g., random intercept but fixed slopes)
- two-stage approach to estimate detection probability? (first estimate detection probability, then use estimates in regression model)
- do not necessarily estimate non-linear trend over multiple years, but also just difference between subsequent years (initial population size matters)


### Implementation

Some thoughts:

- rely as much as possible on existing implementations 
- options
  - R packages for 'N-mixture' modeling (however: no random effects?)
  - R packages for generalized linear mixed effects modeling (freq./Bayes; however: no latent outcomes?)
  - R packages for generalized linear SEM?
  - existing custom implementation in e.g. stan?

## Simulation study


### Design

Some thoughts:

- number of sampling locations / sample size: vary 
- days of recording per location: vary optionally, keep fixed initially
- if both are varied, possibility to look at trade-off effects (in terms of power, but also costs)

::: {.callout-note icon=false}
### ?
Where to get plausible population-level parameter values from?

- Analysis of existing data?
- Expert judgement?
:::


### Implementation

- again, make use of helpers:
  - simr
  - mlpwr (to later include budget constraints or cost trade-offs between design parameters)
  - simDesign




## References

::: {#refs}
:::
