---
title: "Power analyses for chinese muntjac"
format: 
  html:
    toc: true
    toc-location: right-body
    toc-depth: 3
    toc-expand: true
    theme: yeti
    number-sections: true
    number-depth: 3
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
lightbox: true
bibliography: ../../../media/MIAS.bib
execute: 
  eval: true
  echo: true
  cache: false
---

```{r setup}




```


<!----------------------------------------------------------------------------------------------------------------------------------------------->
<!-- Overview ----------------------------------------------------------------------------------------------------------------------------------->
<!----------------------------------------------------------------------------------------------------------------------------------------------->

# Overview

The present documents illustrates simulation approaches for the following **monitoring scopes**:

- A: Abundance under management - status
  - Determine population size under sustained eradication measures in distribution area in province of Antwerp
  - ...
- B: Abundance under management - trend
  - Estimate changes in population size in relation to sustained eradication measures in above distribution area
  - Can concern changes in absolute or relative abundance, depending on the assumptions we are willing to make
  - Will reveal a minimal sample size required for detecting a certain effect with a certain statistical power, 
  or minimal effects that can be detected with certain sample sizes
- C: Detection under management - status
  - Detect potential spread into buffer area around above distribution area in relation to sustained eradication measures in distribution area
- D: Detection after management - status
  - Detect potential left-over animals in above distribution area after eradication has been completed 


For these scopes, the document summarizes some thoughts on the following components of a simulation-based power analysis:

- Sampling
  - How do (simulated) data map onto geographical locations?
- Power / precision analysis
  - What is an appropriate data-generating model?
  - Thoughts on a possible analysis setup

<!----------------------------------------------------------------------------------------------------------------------------------------------->
<!-- Sampling ----------------------------------------------------------------------------------------------------------------------------------->
<!----------------------------------------------------------------------------------------------------------------------------------------------->
# Sampling

Thinking about sampling is useful as it answers the question: 
How do (simulated) data map onto geographical locations?

This

  - is needed to visualize the data collection-side under different scenario's
  - helps judging whether assumptions during data generation are more or less realistic
  - helps determining the range of realistic sample sizes

The following sections in this chapter deal with

- the definition of a sample frame
- the drawing of a random sample of locations
- the mapping of future simulated data on the random sample of locations


## Define the sample frame

To define the sample frame, we take the following 2 preparatory steps:

1. Get map layers: 
  - Flanders' municipalities (as distribution is currently tied to specific municipalities)
  - Flanders' habitat area (e.g., forest)


```{r sampling-get-map-layers}
#| include: true
#| results: 'hide'

# --- get map layers---------------------------------------------

# municipalities
map_mun <- inbospatial::get_feature_wfs(
  wfs = "https://geo.api.vlaanderen.be/VRBG/wfs",
  layer = "VRBG:Refgem"
)
# https://www.vlaanderen.be/datavindplaats/catalogus/
# voorlopig-referentiebestand-gemeentegrenzen-
# toestand-01-01-2025-correctie-11-02-2025
#
# check CRS
if (FALSE) sf::st_crs(map_mun)$Name # EPSG:31370

# habitat
map_hab <- sf::st_read("../../../data/gis/inbo_boskartering/bosvl2001.shp")
# get forest map (vector?)
# raster: https://www.vlaanderen.be/datavindplaats/catalogus/
# digitale-boswijzer-vlaanderen-2021
# raster: https://www.vlaanderen.be/datavindplaats/catalogus/
# groenkaart-vlaanderen-2021
#
# check CRS
if (FALSE) sf::st_crs(map_hab)$Name # EPSG:31370


```

2. Define relevant areas:
  - Get distribution area by selecting the relevant municipalities: Schoten, Brasschaat, Wuustwezel
  - Get buffer area of width $u$ m around distribution area (here $u = 2000$ m)
  - Get habitat regions within distribution area (e.g., Forests)
  - Get habitat regions within buffer area

```{r sampling-process-map-layers}
#| include: true
#| cache: true
#| results: 'hide'
#| warning: false

# TO DO: check warnings

# --- process map layers ---------------------------------------------

# select relevant municipalities
map_mun_antw <- map_mun |>
  dplyr::filter(
    grepl("Schoten|Brasschaat|Wuustwezel", NAAM)
  )

# extract polygons from multisurface to enable further processing
# (mapview and intersection)
lapply(1:nrow(map_mun_antw),
       function (i) {
         assertthat::are_equal(
           sf::st_as_text(map_mun_antw$SHAPE[i]) |> grep("POLYGON", x = _),
           1
         )
         })
map_mun_antw <- sf::st_cast(map_mun_antw, "GEOMETRYCOLLECTION") |>
  sf::st_collection_extract("POLYGON")
if (FALSE) mapview::mapview(map_mun_antw)

# create buffer around municipalities
u = 2000
map_mun_antw_merged <- map_mun_antw |>
  dplyr::summarise()
tmp_buff <- map_mun_antw_merged |>
  sf::st_buffer(
    dist = u # meters, since Belgian Lambert
    )
map_mun_antw_buff <- sf::st_difference(
  tmp_buff,
  map_mun_antw_merged
)
if (FALSE) mapview::mapview(map_mun_antw_buff)

# quickly check municipality maps
if (FALSE) {
  mapview::mapview(
    list(map_mun_antw, map_mun_antw_buff),
    col.regions = list("blue", "red"),
    alpha.regions = .3
  )
}

# intersect habitat with municipality maps
map_hab <- sf::st_transform(x = map_hab, crs = "EPSG:31370")
map_mun_antw_hab <- sf::st_intersection(
  map_mun_antw,
  map_hab
  )
mapview::mapview(map_mun_antw_hab)
map_mun_antw_buff_hab <- sf::st_intersection(
  map_mun_antw_buff,
  map_hab
)
if (FALSE) mapview::mapview(map_mun_antw_buff_hab)

```

After these 2 steps we get the following map.

```{r fig-plot-dist-buff-area}
#| include: true
#| cache: true
#| results: 'hide'
#| message: false
#| fig-cap: "Map of distribution and surrounding buffer area (both in yellow) and enclosed habitat regions (both in green). The distribution area is highlighted by lighter colors and a solid contour." 
#| out-width: 150%
#| layout-ncol: 1 

# transform: pseudo-mercator, which is CRS of map tile 
# (unless map tile - in pseudo-mercator - is first layer)
mapnames <-   c(
    "map_mun_antw",
    "map_mun_antw_hab",
    "map_mun_antw_buff",
    "map_mun_antw_buff_hab",
    "map_mun_antw_merged"
  )
maplist <- lapply(
  mapnames,
  \(i){
    tmpmap <- sf::st_transform(x = get(i), crs = "EPSG:3857")
  }
) 
names(maplist) <- paste0(mapnames, "_3857")
# check transformation
lapply(maplist, \(x) sf::st_crs(x)$Name)

# base plot
plot_base <- ggplot2::ggplot(
  data = maplist$map_mun_antw_3857 # warning if wgs84 is used
) +
  ggspatial::annotation_map_tile(
    type = "cartolight", zoom = 12, cachedir = tempdir(), alpha = .5 
  ) 
plot_names <- ggplot2::geom_sf_text(
    ggplot2::aes(
      label = NAAM
    ),
    size = 5,
    color = INBOtheme::vl_black,
    check_overlap = TRUE
  )

# plot distribution area
geoms_dist <- list(
  ggplot2::geom_sf(fill = INBOtheme::vl_yellow, 
                   alpha = .6),
  ggplot2::geom_sf(data = maplist$map_mun_antw_hab_3857, 
                   fill = INBOtheme::vl_lightgreen, 
                   color = INBOtheme::vl_lightgreen,
                   alpha = .6) 
)
plot_dist <- plot_base + geoms_dist + plot_names

# plot buffer area
geoms_buffer <- list(
    ggplot2::geom_sf(data = maplist$map_mun_antw_buff_3857, 
                     fill = INBOtheme::vl_darkyellow, 
                     alpha = .6),
  ggplot2::geom_sf(data = maplist$map_mun_antw_buff_hab_3857, 
                   fill = INBOtheme::vl_darkgreen, 
                   color = INBOtheme::vl_darkgreen,
                   alpha = .6)
)
plot_buff <- plot_base + geoms_buffer + plot_names

# combined plot
plot_dist_buff <- plot_base + geoms_dist + geoms_buffer +   plot_names +
  # emphasize dist
  ggplot2::geom_sf(data = maplist$map_mun_antw_merged_3857, 
                   fill = NA, 
                   color = INBOtheme::vl_black, 
                   linewidth = 0.8) 
plot_dist_buff

```

On the basis of these, we define the sample frame (NL: *steekproefkader*) for the monitoring scopes targeting the distribution (and at the same time management) area, i.e., scopes A, B, and D, as the habitat regions within the distribution area.

The sample frame for monitoring scope C, which targets a buffer area around the distribution area is defined as the habitat regions within the buffer area. 


::: {.callout-note icon=false}
##### Specifications

- Define distribution area
  - which area?
    - [x] known municipalities
    - [ ] most recent (clustered) GBIF data
  - include (smaller) buffer area?
    - [ ] yes, namely ...
    - [x] no
- Define habitat area
  - which land cover/use?
    - [x] forest polygons from *Boskartering 2001*
    - [ ] forest on the basis of OpenStreetMap information (cf. [sample frame partridge](https://zenodo.org/records/8084777))
    - [ ] forest based on [Vito's land use data base](https://geodynamix.eu/land-use-database) 
  - take spatial clustering into account?
    - [ ] yes, namely ...
    - [x] no
- Define buffer area
  - around which area?
    - [x] distribution area as a whole
    - [ ] habitat areas within distribution area
  - which width (i.e., $u$)?
    - [x] arbitrary value for illustration purposes
    - [ ] value based on expected daily range of species
- Define sample frame
  - how to define intersection with distribution/buffer area?
    - [x] habitat regions overlap with distribution/buffer area 
    - [ ] overlap at least $x \gg 0$ %
  
:::

#### Sample frame based on OpenStreetMap information

The following map is comparable to the previous one, but habitat areas are now identified based on current OpenStreetMap data ([category "forest" and "orchard"](https://wiki.openstreetmap.org/wiki/Forest)).


```{r fig-plot-osm}
#| include: true
#| cache: true
#| results: 'hide'
#| message: false
#| fig-cap: "Map based on OpenStreetMap information" 
#| out-width: 150%
#| layout-ncol: 1 

maplist_3857 <- readRDS("../../../data/gis/sample/sampleframe_maplist_3857.Rds")

# base plot
plot_base <- ggplot2::ggplot(data = maplist_3857$map_antw) +
  ggspatial::annotation_map_tile(
    type = "cartolight", zoom = 12, cachedir = tempdir(), alpha = .5
  )
plot_names <- ggplot2::geom_sf_text(ggplot2::aes(label = NAAM),
  size = 5,
  color = INBOtheme::vl_black,
  check_overlap = TRUE
)

# distribution area
geoms_dist <- list(
  ggplot2::geom_sf(fill = INBOtheme::vl_yellow,
                   alpha = .6),
  ggplot2::geom_sf(data = maplist_3857$map_antw_hab,
                   fill = INBOtheme::vl_lightgreen,
                   color = INBOtheme::vl_lightgreen,
                   alpha = .6)
)

# buffer area
geoms_buffer <- list(
  ggplot2::geom_sf(data = maplist_3857$map_buff,
                   fill = INBOtheme::vl_darkyellow,
                   alpha = .6),
  ggplot2::geom_sf(data = maplist_3857$map_buff_hab,
                   fill = INBOtheme::vl_darkgreen,
                   color = INBOtheme::vl_darkgreen,
                   alpha = .6)
)

# combined plot
plot_dist_buff <- plot_base + geoms_dist + geoms_buffer + plot_names +
  # emphasize dist
  ggplot2::geom_sf(data = maplist_3857$map_antw_merged,
                   fill = NA,
                   color = INBOtheme::vl_black,
                   linewidth = 0.8)
plot_dist_buff


```


## Draw a sample

For this illustration, we draw a naive random sample of locations.
To this end, we take the following steps:

1. Tessellation of the sample frame:
  - Overlay distribution / buffer area with spatial grid consisting of quadratic cells of dimension $p \times p$ m [here $100 \times 100$ m for illustrational purposes]
  - Extract grid cells that show $q$ % overlap with habitat regions within distribution area (here $q > 0$) 

```{r sampling-create-grid}
#| include: true
#| cache: true
#| results: 'hide'

# create grid over bounding box around distribution area
p = 100
grid_box <- sf::st_make_grid(
  map_mun_antw_merged,
  cellsize = p # meters, since Belgian Lambert
) |>
  sf::st_sf()
if (FALSE) mapview::mapview(list(map_mun_antw_merged, grid_box))

# extract grid cells that overlap with habitat polygons
grid_hab <- sf::st_filter(
  grid_box,
  map_mun_antw_hab
  #,.predicate = sf::st_within
)
if (FALSE) mapview::mapview(list(map_mun_antw_hab, grid_hab))

# extract grid cells that overlap with habitat polygons
# https://github.com/r-spatial/sf/issues/1125


```

2. Draw random sample of $C$ grid cells extracted in the previous step (here $100$ grid cells)

```{r sampling-draw-random-sample}
#| include: true
#| cache: true
#| results: 'hide'

# --- draw naive random sample ---------------------------------------------

# sample size
ncell <- 500
assertthat::assert_that(ncell < nrow(grid_hab))

# sample
grid_sample <- grid_hab |>
  dplyr::sample_n(tbl = _, size = ncell)

# GRTS
# https://inbo.github.io/grtsdb/articles/basic.html
# implementation?

```

The following plot visualizes the grid and the sample drawn.

```{r fig-plot-sample}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Visualization of the sampled (filled) vs. not sampled (empty) grid cells"
#| out-width: 150%

# transform: pseudo-mercator
grid_hab_3857 <- sf::st_transform(x = grid_hab, crs = "EPSG:3857")
grid_sample_3857 <- sf::st_transform(x = grid_sample, crs = "EPSG:3857")


# sample from sampling frame distribution area
plot_dist_sample1 <- plot_dist +
  ggplot2::geom_sf(data = grid_hab_3857, color = INBOtheme::vl_lightred, alpha = .3) +
  ggplot2::geom_sf(data = grid_sample_3857, color = NA, fill = INBOtheme::vl_lightred) +
  plot_names

plot_dist_sample1

```


Within each sampled grid cell, a camera trap is assumed to be placed following a certain strategy (e.g., placement in center or corner of grid cell). 

The total number of grid cells associated with the sample frame forms the upper bound of sample sizes to consider in a power / precision analysis.


::: {.callout-note icon=false}
##### Specifications / Extensions

- Determine grid
  - cell size?
    - [ ] value from literature (250 m; cf. @bollen_gebruik_nodate)
    - [ ] value based on camera angle/radius (usually $< 10$ m?)
    - [x] smallish value to keep things visible
    - [ ] value based on habitat parameters (e.g., forest density; to make sure a grid cell contains suitable areas for camera placement)
  - how to define intersection with habitat areas?
    - [x] grid cell overlaps with habitat areas 
    - [ ] overlap at least $q \gg 0$ %
- Draw a random sample
  - strategy?
    - [x] naive random sample
    - [ ] spatially balanced random sample via the GRTS method (INBO R package *grtsdb*)

:::




<!----------------------------------------------------------------------------------------------------------------------------------------------->
<!-- Simulation ---------------------------------------------------------------------------------------------------------------------------->
<!----------------------------------------------------------------------------------------------------------------------------------------------->

# Power / precision analysis

## Some general remarks

For the variables occurring in any of the following data-generating / analysis models we can distinguish the following dimensions:

- $i = 1, ..., I$ sampling locations
- $d = 1, ..., D$ days per measurement period (more precisely: $d_{t}$; here the same for all $t$, so subscript $t$ is left out)
- $t = 1, ..., T$ years, with one measurement period per year

Note that we here do not consider the case of multiple measurement periods per year.
In that case, differences between periods within years could be allowed in the model, capturing seasonal deviations from a longer-term (i.e., year-level) trend (e.g., more individuals before than after winter).
For now, we however simply assume that the optimal period per year is used to obtain measurements.
Optimality is for instance tied to a maximal detection probability, the absence of complicating factors such as ongoing control actions during measurement and so on.

We can think about possible data-generating models in the context of the generalized linear (mixed) modeling framework.
Sticking to the not-too-complex model variants, we can rely on well-known properties/behavior and existing implementations for both data generation and model estimation. 
(Whether the model is feasible to implement as a data-analysis model should be considered already at this stage to not complicate things later.)
We do this for abundances.
For presences, we revert to a simpler option.
For now we generally assume that we use the same model for data generation as we use for data-analysis.


## Monitoring scopes A & B 

### Data-generating model

#### Data-generating distribution

##### Poisson distribution

Monitoring scopes A and B (abundance under management - status & trend) concern abundances of animals as outcomes (i.e., numbers of individuals per location), so we will rely on the **Poisson distribution** for these cases, with the parameter of focus being the expected number of observed animals:

$$
nobs_{i,d,t} \sim Pois(\mu_{i,t})
$$ {#eq-pois}
with

- $nobs_{i,d,t} \in \{0,1,2,3, ...\}$: number of observations of animals at location $i$, day $d$, year $t$
- $\mu_{i,t} \in (0, \infty)$: expected number of observations of animals at location $i$, year $t$ 
  - within a location and year, measurements from different days are treated as replicates, all independent realizations from the same Poisson distribution 
  - this reflects assumption of some stability over time (the alternative would be to assume variations across days and have day enter a regression model in some way (e.g., random effect, temporal trend of some shape))

When looking at status, we can assume we are looking at one specific year, so $t = t'$.

In the case of @eq-pois, we assume that $\mu_{i,t}$ is linked to the expected true population size / the true abundance at location $i$, year $t$, $\nu_{i,t}$.^[Following the logic of N-mixture modeling, or latent variable modeling, one would rather assume that $nobs_{i,d,t}$ is linked to the true population size at location $i$, year $t$, $n_{i,t}$; if the link holds at that level then it should also be implied at the level of expectations, but not the reverse.]
For simplicity, we assume a linear relationship that is invariant across all dimensions considered (i.e., space and time), but also across abundance itself:

$$
\mu_{i,t} = a * \nu_{i,t}
$$ {#eq-pois2}

In an ideal case $a = 1$, which means we can make statements about absolute abundances over space, time and abundance.
If $a \neq 1$, then we can still make statements about relative abundances over space, time and abundance *without having to know* $a$.
Reasons for $a \neq 1$ can be: an imperfect detection probability; animals being recorded more than once within a day. 

However, this assumption might be problematic if animals are very territorial, with no or little overlap between territories.
Then we are most likely only observing one to a few individuals and the number of observations is more related to their activity during that day.
In that case, we would need to find another approach to generate and analyse abundances.
One option could be the logistic regression model (see notes later), if abundances are basically one at this geographical scale, then we are dealing with data of presence and absence.

##### Bernoulli distribution

If abundance at the given geographical scale is rather about the presence of animals, then we can alternatively rely on the **Bernoulli distribution** as the data-generating distribution. The parameter of focus then is the probability of presence:

$$
obs_{i,d,t} \sim Bern(\psi_{i,t})
$$ {#eq-bern}

- $obs_{i,d,t}  \in \{0,1\}$: observation of animals at location $i$, day $d$, year $t$
- $\psi_{i,t}  \in [0,1]$: probability of presence at location $i$, year $t$
  - within a location and year, measurements from different days are again treated as independent realizations from one Bernoulli distribution 
  - this again reflects assumption of some stability over time

In analogy to @eq-pois2, we can here think of the probability of presence featuring in @eq-bern as being an observed one, which, again, 
should be linked to a true probability of presence.
Ideally the link is a one-to-one link.
Again, an imperfect detection probability can weaken the link.
The same animals being present multiple times or multiple animals being present should not be an issue and can be accommodated by thinking of $\psi$ as the probability of at least one animal being present.


#### Regression model under monitoring scope B

##### Poisson regression

In the case of both monitoring scopes A & B (abundance under management - status & trend), we can rely on the **Poisson regression model**. Starting from the distributional assumption in @eq-pois, we linearly decompose $\mu_{i,t}$, the parameter of the Poisson distribution, after applying the log-link function:

$$
log(\mu_{i,t}) = \eta_{i,t} \\
\Leftrightarrow \\
\mu_{i,t} = exp(\eta_{i,t})
$$ {#eq-poisreg1}
with:

- $log(\cdot)$: natural logarithm, here the log-link function
- $\eta_{i,t}$: linear predictor, which is some linear function of a set of predictor variables
- $exp(\cdot)$: exponential (i.e., inverse of log-link) function

The most simple model variant under monitoring scope B (abundance under management - trend) includes an intercept and a linear effect of year, both location unspecific, so index $i$ is dropped:

$$
\eta_{t} = \beta_{0} + \beta_{1} \, year_t^*
$$ {#eq-poisreg2}
with:

- $\beta_{0}$: intercept; linear predictor if $year_t^*= 0$
- $\beta_{1}$: slope associated with $year_t*$, which is some linear function of year

The **parameter of interest** - corresponding to our monitoring target 'trend' - is obviously the **slope**, $\beta_{1}$.
It's interpretation in terms of the linear predictor, $\eta$, is the one we are familiar with from simple linear regression:
$\beta_1$ reflects the absolute change in linear predictor values given a one-unit change in year*.

The interpretation in terms of the expected number of observed animals, $\mu$, needs to incorporate the link-function used:
Here, $exp(\beta_1)$ reflects the relative difference (ratio) in expected numbers of observations given a 1-unit difference in year*.
We can derive this as (leaving out subscript $t$):

\begin{align}
log(\mu)|(year^*_1 + 1) - log(\mu)|year^*_1 &= \beta_1 \quad\quad|\;exp(\cdot) \\
exp(log(\mu)|(year^*_1 + 1) - log(\mu)|year^*_1) &= exp(\beta_1) \\
\frac{exp(log(\mu)|(year^*_1 + 1))}{exp(log(\mu)|year^*_1)} &= exp(\beta_1) \\
\frac{\mu|(year^*_1 + 1))}{\mu|year^*_1} &= exp(\beta_1)
\end{align}

So, $exp(\beta_1)$ is the factor by which $\mu$ changes given a one-unit change in year
($\mu|(year^*_1 + 1) = exp(\beta_1)*\mu|year^*_1$),
and $(exp(\beta_1)-1)*100$ is the percentage change in $\mu$ associated with a one-unit change in year.

The following figure shows a number of different trends implied by different sizes of $\beta_1$ for both $\nu$ and $\mu$.

```{r fig-size-trend}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Visualization of different effect sizes of beta_1"
#| out-width: 150%

data_trend <- data.frame(
  # size of linear effect of year
  effect_size = -seq(0, 2, 0.2)
) |>
  tidyr::crossing(
   year_star = seq(0,9)
) |>
  dplyr::mutate(
    eta =  effect_size*year_star + 2,
    mu = exp(eta),
    effect_size = factor(effect_size) |> forcats::fct_rev()
    ) |>
  tidyr::pivot_longer(
    cols = tidyselect::all_of(c("eta", "mu")),
    names_to = "out_name",
    values_to = "out_value"
  )
ggplot2::ggplot(
  data = data_trend, 
  mapping = ggplot2::aes(x = year_star, y = out_value, group = effect_size)
  ) + 
  ggplot2::geom_line(ggplot2::aes(color = effect_size), linewidth = 1.2) +
  ggplot2::facet_wrap(~ out_name, scales = "free_y", ncol = 2) + 
  ggplot2::labs(x = "year*", y = "outcome", color = "size beta1") +
  ggplot2::scale_x_continuous(breaks = data_trend$year_star |> unique()) + 
  ggplot2::scale_colour_viridis_d(direction = -1) + 
  ggplot2::theme_bw()

```

In case $\beta_1 = 0$, we have `r round((exp(0)-1)*100, digits = 0)` %,
if $\beta_1 = -0.2$, we have around `r round((exp(-0.2)-1)*100, digits = 0)` %, 
if $\beta_1 = -0.4$, we have around `r round((exp(-0.4)-1)*100, digits = 0)` %, 
if $\beta_1 = -1$, we have around `r round((exp(-1)-1)*100, digits = 0)` % decrease per year. 

What could be an interesting effect size from a biological perspective?

##### Logistic regression

Alternatively, we can start from the distributional assumption made in @eq-bern and rely on the **Logistic regression model** 
and we linearly decompose $\psi_{i,t}$, the parameter of the Bernoulli distribution, after applying the logit-link function:


$$
logit(\psi_{i,t}) = log(\frac{\psi_{i,t}}{1-\psi_{i,t}}) = \eta_{i,t} \\
$$ {#eq-logreg1}
with:

- $logit(\cdot)$: logit-link function (i.e., the log-odds/ logarithm of the odds)
- $log(\cdot)$: natural logarithm
- $\frac{\psi_{i,t}}{1-\psi_{i,t}}$: odds of presence
- $\eta_{i,t}$: linear predictor

The most simple model variant is the one from above (@eq-poisreg2), with the parameter of interest again being the **slope**, $\beta_{1}$.
This parameter can again be interpreted in terms of the linear predictor,
in terms of the odds of presence ($odds_{i,t}: exp(\eta_{i,t}) = \frac{\psi_{i,t}}{1-\psi_{i,t}}$),
or in terms of the probability of presence ($\psi_{i,t} = \frac{odds_{i,t}}{1+odds_{i,t}}$)

The interpretation in terms of the odds of presence is analoguous to the interpretation above:
So, $exp(\beta_1)$ is the factor by which the odds of presence change given a one-unit change in year
The following figure shows a number of different trends implied by different sizes of $\beta_1$ for $\nu$ (the log-odds), the odds and $\psi$.

```{r fig-size-trend-2}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Visualization of different effect sizes of beta_1 in the logistic regression model"
#| out-width: 150%

data_trend <- data.frame(
  # size of linear effect of year
  effect_size = -seq(0, 2, 0.2)
) |>
  tidyr::crossing(
   year_star = seq(0,15)
) |>
  dplyr::mutate(
    eta =  effect_size*year_star + 2,
    odds = exp(eta),
    psi = odds / (1 + odds),
    effect_size = factor(effect_size) |> forcats::fct_rev()
    ) |>
  tidyr::pivot_longer(
    cols = tidyselect::all_of(c("eta", "odds", "psi")),
    names_to = "out_name",
    values_to = "out_value"
  )
ggplot2::ggplot(
  data = data_trend, 
  mapping = ggplot2::aes(x = year_star, y = out_value, group = effect_size)
  ) + 
  ggplot2::geom_line(ggplot2::aes(color = effect_size), linewidth = 1.2) +
  ggplot2::facet_wrap(~ out_name, scales = "free_y", ncol = 3) + 
  ggplot2::labs(x = "year*", y = "outcome", color = "size beta1") +
  ggplot2::scale_x_continuous(breaks = data_trend$year_star |> unique()) + 
  ggplot2::scale_colour_viridis_d(direction = -1) + 
  ggplot2::theme_bw()

```


If $\beta_1 = -0.2$, then the odds change by a factor of `r round(exp(-0.2), digits = 2)` per year, 
and if $\beta_1 = -1$, that factor is `r round(exp(-1), digits = 2)`. 

The question is again, what could be an interesting effect size from a biological perspective?

#### Regression model under monitoring scope A 

For monitoring scope A (abundance under management - status), we could consider a special case of the model specified for monitoring scope B. 
Specifically, we could further simplify the above regression model (@eq-6) to an intercept-only model to estimate expected local population size at year $t=t'$.
The following thoughts on analysis setup for monitoring scope B then apply analogously to scope A.

Alternatively, we might instead of a somewhat artificial regression model just use a Poisson test (see later elaborations on the Binomial test).

### Analysis setup

Based on the simple model specified above (@eq-6), we can now either

- determine a few minimal trend sizes of interest, and analyse how statistical power to detect those effects varies with the number of locations 
- determine a few interesting numbers of locations, and analyse how the size of the minimal effect that can be detected with a certain statistical power varies

In addition, we can also extend the above data-generation model in a number of ways, the most obvious ones being

- the inclusion of location-specific effects (e.g., a random intercept and/or a random slope) 
- the accommodation of a non-perfect detection probability

Such additional complexities could be included in both the data-generating and analysis model (case 1),
only the data-generating model (case 2),
only the analysis model (case 3, which we discard as not realistic),
or in none of the models (case 4, which we are in now).
In case 1, we expect that simulating and modeling additional complexities will lead to additional estimation variance,
which will lead to larger sample sizes being required to keep statistical power at the desired level.
In case 2, we expect that simulating but not modeling additional complexities will most often lead to estimation bias, which will also affect power, albeit in less predictable ways. 
Investigating the effects of missspecification is also a bit of a different question and should maybe not be of central interest for the present power analysis (or maybe it should).
However, for a relevant special case, neglected imperfect detection probability, we can form an expectation:
Here a trend will like be underestimated lowering statistical power to detect it.
In general, analyses under cases 1 and 2 will reflect more realistic scenarios, but the simple case 4 can function as a lower bound estimate (i.e., it probably won't get better than this). 

Another direction for extensions concerns sample size as multi-dimensional:
Until now it was implicit that we assumed the number of days to be fixed (e.g., at 30; cf. @bollen_gebruik_nodate), while they could also be varied.
In the simple model above, we should see a one-to-one trade-off effect of days and locations (adding one day at one location should have the same effect as adding one location for one day), which is why we can keep days as well fixed.
This situation changes when costs are considered, where different dimensions can have very different contributions to the overall costs (e.g., leaving an existing camera trap in the field for some more time is cheaper than placing an additional new one).
So, if costs become an additional optimization criterion next to statistical power, then it would make sense to vary sample size along multiple dimensions to investigate trade-offs.
(Also when location-specific effects and/or detection probabilities are estimated, sample size dimension trade-off effects will become more interesting and should probably be added to the picture.) 


## Monitoring scopes C & D

### Data-generating model

#### Data-generating distributions

Monitoring scopes C and D (detection under & after management - status) concern presence of animals as outcomes. Here we have 2 options.
If we consider presence per location, we can rely on the **Bernoulli distribution**, with the parameter of focus being the probability of presence:

$$
obs_{i,d,t=t'} \sim Bern(\psi_{i,t=t'})
$$ {#eq-3}


- $obs_{i,d,t=t'}  \in \{0,1\}$: observation of animals at location $i$, day $d$, year $t'$
- $\psi_{i,d,t=t'}  \in [0,1]$: probability of presence at location $i$, year $t'$
  - within a location and year, measurements from different days are again treated as independent realizations from one Bernoulli distribution 
 - this again reflects assumption of some stability over time

Alternatively, we can focus on the number of locations at which animal might be present, relying on the **Binomial distribution** and, again, the probability of presence:

$$
nobs_{d,t=t'} \sim Binom(\psi_{t=t'}, I)
$$ {#eq-4}


- $nobs_{d,t=t'}  \in \{0,1,2,3, ...\}$: number of observations of animals summed across all locations, at day $d$, year $t'$
- $\psi_{t=t'}  \in [0,1]$: probability of presence at time $t = t'$, which is assumed invariant across all locations
  - within a year, measurements from different days are again treated as independent realizations from the same Binomial distribution
- $I$: the total number of locations

The first angle (@eq-3) makes sense if we want draw on the logistic regression model to include location-specific effects (and spatial dependencies) in our analysis.
The second angle (@eq-4) is more of an aggregate approach.
We discuss the potential benefits of including location-specific effects later on.

In analogy to @eq-2, we can here think of the probability of presence featuring in both @eq-3 and @eq-4 as being an observed one, which, again, 
should be linked to a true probability of presence.
Ideally the link is a one-to-one link.
Again, an imperfect detection probability can weaken the link.
The same animals being present multiple times or multiple animals being present should not be an issue and can be accommodated by thinking of $\psi$ as the probability of at least one animal being present.

Under monitoring scopes C & D (detection under & after management - status) we start even simpler than under scope B:
The data-generating model simply is the Binomial distribution presented in @eq-4.
Two analysis options are sketched in the following section.
These are based on existing work of Thierry's.

### Analysis setup

#### Scenario 1: Focus on the minimal detectable deviation from a target

Starting from a specific target value for the probability of presence (say 5% is some sort of critical threshold from an ecological perspective), we can investigate how the minimal deviation detectable with a desired statistical power changes with the sample size.

We do this in the context of the binomial test.^[Note that we analyse one date per location. These could be from a single day, or aggregated across days. The probability of presence has different interpretations in both cases: In the first case, it is the probability that at least one animal was present at at least one moment during the recording; in the second case, it is the probability that at least one animal was present at at least one of the days. It is important to know for which we want to set a critical value. If it is the first probability, then the number of days should be added into the analysis.]
The binomial test can be used to determine whether a proportion of successes (sightings) in a number of Bernoulli trials (sampling locations) differs from an expected proportion of successes (critical value). 
The test statistic is the binomial distribution (@eq-4).

The following figure visualizes for 100 trials (sampling locations) how much statistical power we have to detect deviations of two different sizes from a certain target (h_a1 and h_a2, relative to h_0).


```{r fig-plot-binomial-power}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Power of the binomial test (I = 100)"
#| out-width: 150%

source("../../_functions/plot_binom_power.R")
plot_binom_power(
    n = 100, # number of trials
    psi_0 = 0.05, # probability of succes under h_0
    psi_a = c(0.1, 0.15), # probability of succes under h_a, or vector if multiple values
    alpha = 0.1 # type 1 error
)$plot_power
```

<!---why is curve smooth for psi_0 = 0? ---->

We can now do this for a whole range of sampling locations and each time determine the deviation that is detected with (at least) the desired power.
This is the minimal detectable deviation given the desired power.
The following figure illustrates how this minimally detectable deviation decreases with sample size for two different target values and testing strategies (one- and two-sided). 


```{r fig-minimal-deviation}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Minimal detectable deviation from target"
#| out-width: 150%

source("../power_binomial.R")

plot_power + 
  ggplot2::labs(
    title = NULL
  )

```


We can further complement this by taking estimation precision into account:
For each sample size, we can for all possible sample configurations (all possible numbers of sightings) obtain the observable effects (probability of presence estimates), especially their confidence interval (CI) estimates.
Based on width and position of the CIs with respect to the target and the minimally detectable effect, we can categorize each observable effect according to the [effectclass-logic](https://inbo.github.io/effectclass/articles/visualisation.html#format_ci-and-classification).
The following figure illustrates this and shows which interpretations we can afford given different sample sizes and probabilities of success (figure requires some adjustments).

<!--- fix threshold issue ---->
<!--- use points ---->


```{r fig-classification}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Classification of observable effects with respect to target and minimal detectable effect"
#| out-width: 150%

plot_classification + 
  ggplot2::labs(
    title = NULL

**Note**: We currently analyse one date per location. These could be from a single day, or aggregated across days. The probability of presence has different interpretations in both cases: In the first case, it is the probability that at least one animal was present at at least one moment during a daily recording session; in the second case, it is the probability that at least one animal was present at at least one of the days. It is important to know for which we want to set a critical value. If it is the first probability, then we can link this to the second probability and achieve higher power by increasing the number of days.
The relationship is $P(at\,least\,one\,sighting\,across\,days) = 1 - P(absence)^{\# days}$ with $P(absence) = 1-P(presence)$.
The following figures shows the relationship between $P(at\,least\,one\,sighting\,across\,days)$ and the number of days for different values of $P(presence)$.

```{r fig-probs}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Probability of at least one sighting across days"
#| out-width: 150%

data_test <- data.frame(
  p_pr = c(0.01, .05, .1, .15, .20)
) |> 
  tidyr::crossing(
    day = seq(1,100)
  ) |>
  dplyr::mutate(
    p_ab = 1 - p_pr,
    p_pr1 = 1 - p_ab^day,
    p_pr = factor(p_pr) |> forcats::fct_rev()
  )
ggplot2::ggplot(
  data = data_test, 
  mapping = ggplot2::aes(
    x = day,
    y = p_pr1,
    group = p_pr,
    color = p_pr
  )
) + 
  ggplot2::geom_line(linewidth = 1.2) +
  ggplot2::labs(x = "observation days", y = "probability of at least one sighting", color = "probability of presence") +
  ggplot2::scale_colour_viridis_d(direction = 1) + 
  ggplot2::theme_bw()
  

```


#### Scenario 2: Focus on case of zero sightings

<!---also link to power? ---->

The alternative is to just consider the case in which we observe zero animals:
If at least one observation of an animal is made, action has to be taken anyway.
If zero observations are made across locations, however, we can guess that there are no animals
(i.e., the (maximum likelihood) point estimate of the probability of presence is zero).
The precision associated with this guess however depends on the number of sampling locations.
Intuitively, we will trust our guess more having 100 locations without sighting as compared to 10 locations without sighting.
The more locations without sightings, the less likely/plausible probabilities of presence estimates larger than zero become.
We can formalize this by looking at the confidence interval estimate we get for increasing numbers of zero sightings.
For 72 locations, the CI estimate excludes a probability of presence of 5%, which might be a critical threshold from an ecological perspective. 
A probability of presence of 1% is however still included in the interval. 
We desire the sample size for which the interval does not include a minimal critical value (which needs to be defined against the relevant ecological background) / does only include uncritical values.^[Interpreting single confidence interval estimates is tricky, and I need to double check whether this interpretation is admissible (enough); for now based on Lakens:https://lakens.github.io/statistical_inferences/07-CI.html]


```{r fig-plot-binomial-ci}
#| include: true
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Confidence interval binomial test"
#| out-width: 150%

source("../../_functions/sim_binom_ci.R")

tmp <- sim_binom_ci(
    n_max = 500, # max number of trials
    n_min = 0, # min number of trials
    successes = 0, # number of successes
    psi_0 = 0, # probability of success under h_0 - does not matter for ci
    alternative = "two.sided", # does matter for CI,
    alpha = 0.05 
)
tmp$plot_ci + 
  ggplot2::labs(
    x = "Number of sampling locations",
    y = "Probability of presence"
  )

```


**Note**: The left-hand panel of @fig-classification and @fig-plot-binomial-ci look quite similar.
In fact, the displayed values are pretty close.^[They are almost exactly the same, when the CI is estimated in the context of a one-sided test, but it looks like the implementation is not entirely correct at this point as the estimated upper value does not match the value obtained from the closed-form expression known for this case (see https://en.wikipedia.org/wiki/Binomial_proportion_confidence_interval#Clopper%E2%80%93Pearson_interval)].

Relying on the inverse relationship between a CI and a statistical hypothesis test (ref: Casella & Berger),
we know that we would have rejected the null hypothesis of a zero probability of presence had we observed presences expected under probabilities not included in the CI ($\alpha =$ `r args_ci$alpha`; verify interpretation).
This per se does not guarantee that we end up with the desired power (here `r 1 - args_ci$alpha`).
In the case of psi = 0 however, we come close to the desired power once we reject the null hypothesis.
Intuitively, this probably has in some way to do with things being bounded at 0, but I got stuck and stopped looking into this.
(Knowing that the CI always lies entirely to the right of 0, we would need to show that here $\alpha \approx \beta$, so that $(1 - \alpha) \approx (1- \beta)$).



```{r fig-plot-binomial-power-2}
#| include: false
#| cache: true
#| message: false
#| results: 'hide'
#| fig-cap: "Power of the binomial test (I = 10)"
#| out-width: 150%


args_ci <- list(
    n_max = 100, # max number of trials
    n_min = 0, # min number of trials
    successes = 0, # number of successes
    psi_0 = 0, # probability of success under h_0 - does not matter for ci
    alternative = "less", # for checking interpretation
    alpha = 0.1 
)
tmp <- do.call(sim_binom_ci, args_ci)

test <- tmp$data_ci |> 
  dplyr::filter(
    n == 2,
    grepl("ci_u", ci_name)
  ) 

plot_binom_power(
    n = test$n, # number of trials
    psi_0 = 0, # probability of succes under h_0
    psi_a = test$ci_value, # probability of succes under h_a, or vector if multiple values
    alpha = test$alpha # type 1 error
)$plot_power
```
# References

::: {#refs}
:::
